---
title: "Understanding Spatial MCMC: A Comprehensive Guide"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document provides a comprehensive explanation of the `run_spatial_mcmc` function, which implements a Bayesian spatial linear regression model using Markov Chain Monte Carlo (MCMC) sampling. We'll break down the statistical model, the mathematical foundations, and each step of the MCMC algorithm.

# 1. The Statistical Model

The model assumes:

$$Y = X\beta + W + \varepsilon$$

Where:
- $Y$: Response variable ($n \times 1$)
- $X$: Design matrix ($n \times p$) of covariates
- $\beta$: Regression coefficients ($p \times 1$)
- $W$: Spatial random effects ($n \times 1$) with $W \sim \mathcal{MVN}(0, \sigma^2 R(\phi))$
- $\varepsilon$: Independent errors with $\varepsilon \sim \mathcal{MVN}(0, \tau^2 I)$

## 1.1 Spatial Correlation Functions

The spatial correlation matrix $R(\phi)$ depends on the distance between locations:

### Exponential Correlation:
$$R(\phi)_{ij} = \exp\left(-\frac{d_{ij}}{\phi}\right)$$

### Matérn Correlation (with $\kappa = 1.5$):
$$R(\phi)_{ij} = \left(1 + \frac{d_{ij}}{\phi}\right) \exp\left(-\frac{d_{ij}}{\phi}\right)$$

Where $d_{ij}$ is the distance between locations $i$ and $j$, and $\phi$ is the spatial range parameter.

# 2. Prior Distributions

The function uses conjugate priors:

## Regression Coefficients ($\beta$)
$$\beta \sim \mathcal{MVN}(\beta_{\text{mean}}, \beta_{\text{var}} \cdot I)$$
Where $\beta_{\text{mean}} = 0$ and $\beta_{\text{var}} = 1000$ (vague prior)

## Spatial Variance ($\sigma^2$)
$$\sigma^2 \sim \text{Inverse-Gamma}(\alpha_{\sigma} = 2, \beta_{\sigma} = 1)$$

## Error Variance ($\tau^2$)
$$\tau^2 \sim \text{Inverse-Gamma}(\alpha_{\tau} = 2, \beta_{\tau} = 1)$$

## Spatial Range ($\phi$)
$$\phi \sim \text{Gamma}(\alpha_{\phi} = 2, \beta_{\phi} = 1)$$

# 3. The `rmvn_precision` Function

```{r}
rmvn_precision <- function(precision, linear_term) {
  U <- chol(precision)
  posterior_mean <- backsolve(U, forwardsolve(t(U), linear_term))
  random_part <- backsolve(U, rnorm(length(linear_term)))
  as.numeric(posterior_mean + random_part)
}
```

## 3.1 Mathematical Foundation

This function samples from a multivariate normal distribution parameterized by its **precision matrix** (inverse covariance matrix):

$$X \sim \mathcal{N}(\mu, \Sigma)$$

But instead of providing $\Sigma$ (the covariance matrix), we provide:
- **`precision`**: The precision matrix $Q = \Sigma^{-1}$
- **`linear_term`**: The vector $b = Q\mu$ (precision times mean)

So we're sampling from:

$$X \sim \mathcal{N}(Q^{-1}b, Q^{-1})$$

## 3.2 The Algorithm Steps

### Step 1: Cholesky Decomposition
```{r eval=FALSE}
U <- chol(precision)
```

The Cholesky decomposition factors the precision matrix as:

$$Q = U^T U$$

where $U$ is an upper triangular matrix.

### Step 2: Compute the Posterior Mean
```{r eval=FALSE}
posterior_mean <- backsolve(U, forwardsolve(t(U), linear_term))
```

This solves the system:

$$Q\mu = b \implies U^T U \mu = b$$

### Step 3: Generate Random Noise
```{r eval=FALSE}
random_part <- backsolve(U, rnorm(length(linear_term)))
```

We generate $Z \sim \mathcal{N}(0, I)$ and transform:

$$Y = U^{-1}Z \sim \mathcal{N}(0, Q^{-1})$$

### Step 4: Combine
```{r eval=FALSE}
as.numeric(posterior_mean + random_part)
```

The final sample is:

$$X = \mu + Y \sim \mathcal{N}(\mu, Q^{-1})$$

## 3.3 Mathematical Verification

**Mean**:
$$\mathbb{E}[X] = Q^{-1}b = \mu$$

**Covariance**:
$$\text{Var}(X) = \text{Var}(U^{-1}Z) = U^{-1} \text{Var}(Z)(U^{-1})^T = U^{-1}I(U^{-1})^T = (U^T U)^{-1} = Q^{-1}$$

## 3.4 Why This Approach is Efficient

1. **Computational efficiency**: In spatial models, precision matrices are often sparse
2. **Numerical stability**: Avoids inverting potentially ill-conditioned covariance matrices
3. **Natural for GMRFs**: Many spatial models are naturally expressed in terms of precision matrices

# 4. The MCMC Algorithm - Step by Step

## 4.1 Update $\beta$ (Regression Coefficients)

### Full Conditional Distribution:
$$\beta \mid Y, W, \tau^2 \sim \mathcal{MVN}(\mu_{\beta}, V_{\beta})$$

Where:
$$V_{\beta} = \left(\frac{X^T X}{\tau^2} + P_{\beta}\right)^{-1}$$
$$\mu_{\beta} = V_{\beta} \left(\frac{X^T(Y - W)}{\tau^2} + P_{\beta}\beta_{\text{prior}}\right)$$

Here $P_{\beta} = \text{diag}(1/\beta_{\text{var}})$ is the prior precision matrix.

### Why the Precision Matrix is $X^TX/\tau^2 + P_\beta$

This comes from **Bayesian linear regression theory**. The posterior precision is the sum of:

$$\text{Posterior Precision} = \underbrace{\frac{X^TX}{\tau^2}}_{\text{Data Precision}} + \underbrace{P_{\beta}}_{\text{Prior Precision}}$$

**Intuition**:
- More data ($X^TX$ grows with $n$) = more precision
- Less noise (smaller $\tau^2$) = more precision
- Stronger prior (smaller prior variance) = more precision

### Implementation:
```{r eval=FALSE}
beta_precision <- crossprod(X) / tau2_current + beta_prior_precision
beta_linear_term <- crossprod(X, Y - W_current) / tau2_current + 
                   beta_prior_precision %*% beta_prior_mean
beta_current <- rmvn_precision(beta_precision, beta_linear_term)
```

## 4.2 Update $W$ (Spatial Random Effects)

### Full Conditional Distribution:
$$W \mid Y, \beta, \sigma^2, \tau^2, \phi \sim \mathcal{MVN}(\mu_W, V_W)$$

Where:
$$V_W = \left(\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}\right)^{-1}$$
$$\mu_W = V_W \left(\frac{Y - X\beta}{\tau^2}\right)$$

### Understanding $\sigma^2 R^{-1}$

$\sigma^2 R^{-1}$ represents the **precision matrix** (inverse covariance) of $W$ **scaled by** $\sigma^2$:

$$\text{Precision of } W = \frac{1}{\sigma^2}R^{-1}$$

#### Two Sources of Information:

1. **Data Fit**: $\frac{1}{\tau^2}I$ - How well $W$ explains the residuals
2. **Spatial Prior**: $\frac{1}{\sigma^2}R^{-1}$ - Enforces spatial smoothness

#### What is $R^{-1}$?

For exponential correlation, $R^{-1}$ approximates a **second-order difference operator** (like a Laplacian), which:
- Penalizes deviations from smoothness
- Adjacent locations have negative correlations in the precision matrix
- Enforces smooth spatial patterns

### Implementation:
```{r eval=FALSE}
R_current <- make_correlation(phi_current)
R_inverse <- chol2inv(chol(R_current))
W_precision <- diag(1 / tau2_current, n) + R_inverse / sigma2_current
W_linear_term <- as.numeric(Y - X %*% beta_current) / tau2_current
W_current <- rmvn_precision(W_precision, W_linear_term)
```

### Intuitive Interpretation:

| Scenario | Effect on $W$ |
|----------|---------------|
| Large $\sigma^2$ | Weak spatial prior, $W$ can vary freely |
| Small $\sigma^2$ | Strong spatial prior, $W$ must be smooth |
| Small $\tau^2$ | Data precision high, $W$ fits residuals closely |
| Large $\tau^2$ | Data precision low, $W$ follows spatial prior |

## 4.3 Update $\sigma^2$ (Spatial Variance)

### Full Conditional Distribution:
$$\sigma^2 \mid W, \phi \sim \text{Inverse-Gamma}(\alpha_{\sigma^*}, \beta_{\sigma^*})$$

Where:
$$\alpha_{\sigma^*} = \alpha_{\sigma} + \frac{n}{2}$$
$$\beta_{\sigma^*} = \beta_{\sigma} + \frac{1}{2}W^T R^{-1} W$$

### Implementation:
```{r eval=FALSE}
sigma2_shape <- priors$sigma2_shape + n / 2
sigma2_rate <- priors$sigma2_rate + 0.5 * drop(crossprod(W_current, R_inverse %*% W_current))
sigma2_current <- 1 / rgamma(1, shape = sigma2_shape, rate = sigma2_rate)
```

### Mathematical Note:
The quadratic form $W^T R^{-1} W$ measures how well $W$ fits the spatial correlation structure.

## 4.4 Update $\tau^2$ (Error Variance)

### Full Conditional Distribution:
$$\tau^2 \mid Y, \beta, W \sim \text{Inverse-Gamma}(\alpha_{\tau^*}, \beta_{\tau^*})$$

Where:
$$\alpha_{\tau^*} = \alpha_{\tau} + \frac{n}{2}$$
$$\beta_{\tau^*} = \beta_{\tau} + \frac{1}{2}(Y - X\beta - W)^T(Y - X\beta - W)$$

### Implementation:
```{r eval=FALSE}
residual <- as.numeric(Y - X %*% beta_current - W_current)
tau2_shape <- priors$tau2_shape + n / 2
tau2_rate <- priors$tau2_rate + 0.5 * sum(residual^2)
tau2_current <- 1 / rgamma(1, shape = tau2_shape, rate = tau2_rate)
```

## 4.5 Update $\phi$ (Spatial Range) - Metropolis-Hastings

This is the only non-conjugate update, so we use a Metropolis-Hastings step on $\eta = \log(\phi)$.

### Why $\eta = \log(\phi)$?

#### Problems with Direct $\phi$ Update:
1. **Constraint issue**: $\phi > 0$, normal proposals can give negative values
2. **Boundary effects**: Asymmetric proposal near 0
3. **Scale dependence**: Optimal step size depends on $\phi$'s scale

#### Benefits of Log-Transformation:
1. **Unconstrained parameter space**: $\eta \in (-\infty, \infty)$
2. **Better mixing**: Posterior often more Gaussian-like
3. **Scale-adaptive**: Fixed proposal variance works across scales

### The Proposal:
$$\eta_{\text{current}} = \log(\phi_{\text{current}})$$
$$\eta_{\text{proposed}} \sim \mathcal{N}(\eta_{\text{current}}, \sigma_{\eta}^2)$$
$$\phi_{\text{proposed}} = \exp(\eta_{\text{proposed}})$$

### Implementation:
```{r eval=FALSE}
eta_current <- log(phi_current)
eta_proposed <- rnorm(1, eta_current, phi_proposal_sd)
phi_proposed <- exp(eta_proposed)
```

# 5. Understanding the Metropolis-Hastings Acceptance Ratio for $\phi$

## 5.1 The General Metropolis-Hastings Framework

In Metropolis-Hastings, we accept a proposed value with probability:

$$\alpha = \min\left(1, \frac{\pi(\theta^*) \cdot L(\theta^*) \cdot q(\theta^* \to \theta)}{\pi(\theta) \cdot L(\theta) \cdot q(\theta \to \theta^*)}\right)$$

Where:
- $\pi(\theta)$ = Prior probability of $\theta$
- $L(\theta)$ = Likelihood of data given $\theta$
- $q(\theta \to \theta^*)$ = Proposal distribution (probability of proposing $\theta^*$ given current $\theta$)

## 5.2 Our Specific Case: The Acceptance Ratio

In our code, we're proposing a new value for $\phi$ using:

$$\eta^* = \eta + \text{noise}, \quad \text{where } \text{noise} \sim \mathcal{N}(0, \sigma_{\eta}^2)$$

Since the proposal is symmetric ($q(\eta \to \eta^*) = q(\eta^* \to \eta)$), the proposal ratio cancels out:

$$\alpha = \min\left(1, \frac{\pi(\phi^*) \cdot L(W|\phi^*) \cdot \phi^*}{\pi(\phi) \cdot L(W|\phi) \cdot \phi}\right)$$

Taking logs to avoid numerical underflow:

$$\log(\alpha) = \min\left(0, \log \pi(\phi^*) + \log L(W|\phi^*) + \log(\phi^*) - \log \pi(\phi) - \log L(W|\phi) - \log(\phi)\right)$$

In the code, this is implemented as:

```{r eval=FALSE}
log_target_proposed - log_target_current
```

where:
```{r eval=FALSE}
log_target_current = log π(φ_current) + log L(W|φ_current) + log(φ_current)
log_target_proposed = log π(φ_proposed) + log L(W|φ_proposed) + log(φ_proposed)
```

## 5.3 Breaking Down Each Component

### Component 1: $\log \pi(\phi)$ - The Prior

This is the log of the Gamma prior density:

$$\pi(\phi) = \frac{\beta^{\alpha}}{\Gamma(\alpha)} \phi^{\alpha-1} e^{-\beta\phi}$$

Taking logs:

$$\log \pi(\phi) = \alpha \log(\beta) - \log(\Gamma(\alpha)) + (\alpha-1)\log(\phi) - \beta\phi$$

In the code:
```{r eval=FALSE}
dgamma(phi_current, shape = priors$phi_shape, 
       rate = priors$phi_rate, log = TRUE)
```

**Interpretation:** This penalizes values of $\phi$ that are unlikely under our prior. If $\phi$ is far from the prior mean, this term will be smaller (more negative), making acceptance less likely.

### Component 2: $\log L(W|\phi)$ - The Likelihood

This is the log-likelihood of the spatial random effects $W$ given $\phi$:

$$L(W|\phi) = (2\pi\sigma^2)^{-n/2} |R(\phi)|^{-1/2} \exp\left(-\frac{1}{2\sigma^2} W^T R(\phi)^{-1} W\right)$$

Taking logs:

$$\log L(W|\phi) = -\frac{n}{2}\log(2\pi) - \frac{n}{2}\log(\sigma^2) - \frac{1}{2}\log|R(\phi)| - \frac{1}{2\sigma^2} W^T R(\phi)^{-1} W$$

In the code:
```{r eval=FALSE}
dmvnorm(W_current, mean = rep(0, n), 
        sigma = sigma2_current * R_current, log = TRUE)
```

**Interpretation:** This measures how well the spatial correlation structure (defined by $\phi$) explains the observed spatial patterns in $W$. If $\phi$ makes the spatial structure fit $W$ better, this term will be larger.

### Component 3: $\log(\phi)$ - The Jacobian

This is the Jacobian from transforming $\phi$ to $\eta = \log(\phi)$:

$$\left|\frac{d\phi}{d\eta}\right| = \left|\frac{d}{d\eta}\exp(\eta)\right| = \exp(\eta) = \phi$$

In the code:
```{r eval=FALSE}
eta_current  # This is log(phi_current)
```

**Interpretation:** This accounts for the fact that we're proposing on the log scale. Without this term, we'd be sampling from the wrong distribution because the prior on $\phi$ doesn't equal the prior on $\log(\phi)$.

## 5.4 The Complete Log-Target in Code

```{r eval=FALSE}
log_target_current <- 
    dmvnorm(W_current, mean = rep(0, n), 
           sigma = sigma2_current * R_current, log = TRUE) +  # log L(W|φ)
    dgamma(phi_current, shape = priors$phi_shape,              # log π(φ)
           rate = priors$phi_rate, log = TRUE) +
    eta_current  # log(φ) Jacobian

log_target_proposed <- 
    dmvnorm(W_current, mean = rep(0, n), 
           sigma = sigma2_current * R_proposed, log = TRUE) +  # log L(W|φ*)
    dgamma(phi_proposed, shape = priors$phi_shape,              # log π(φ*)
           rate = priors$phi_rate, log = TRUE) +
    eta_proposed  # log(φ*) Jacobian
```

## 5.5 The Decision Rule

The acceptance ratio determines whether we move to the proposed value:

```{r eval=FALSE}
if (is.finite(log_acceptance_ratio) && log(runif(1)) < log_acceptance_ratio) {
    phi_current <- phi_proposed
    R_current <- R_proposed
    phi_accepted[i] <- TRUE
}
```

### How it works:

1. **Compute log acceptance ratio:**
   $$\Delta = \log \pi(\phi^*) + \log L(W|\phi^*) + \log(\phi^*) - \log \pi(\phi) - \log L(W|\phi) - \log(\phi)$$

2. **Generate random number:** $u \sim \text{Uniform}(0,1)$

3. **Accept if:** $\log(u) < \Delta$ (which is equivalent to $u < \exp(\Delta)$)

### When do we accept?

| Scenario | What happens | Interpretation |
|----------|--------------|----------------|
| $\Delta > 0$ | Always accept | Proposed value has higher posterior density |
| $\Delta < 0$ | Accept with probability $\exp(\Delta)$ | Proposed value has lower posterior density |

## 5.6 A Numerical Example

Let's trace through a concrete example to see how this works:

### Setup:
- Current $\phi = 1.0$, so $\eta = \log(1.0) = 0$
- Proposed $\phi^* = 1.5$, so $\eta^* = \log(1.5) = 0.405$
- Prior: Gamma(shape=2, rate=1)
- Current log-likelihood: -10.0
- Proposed log-likelihood: -8.5

### Step 1: Compute current log-target

```
log_prior_current = dgamma(1.0, shape=2, rate=1, log=TRUE) = -1.0
log_lik_current = -10.0
jacobian_current = log(1.0) = 0
log_target_current = -1.0 + (-10.0) + 0 = -11.0
```

### Step 2: Compute proposed log-target

```
log_prior_proposed = dgamma(1.5, shape=2, rate=1, log=TRUE) = -0.69
log_lik_proposed = -8.5
jacobian_proposed = log(1.5) = 0.405
log_target_proposed = -0.69 + (-8.5) + 0.405 = -8.785
```

### Step 3: Compute acceptance ratio

```
log_acceptance_ratio = -8.785 - (-11.0) = 2.215
acceptance_prob = exp(2.215) = 9.16
alpha = min(1, 9.16) = 1.0
```

**Result:** We always accept because the proposed value has much higher posterior density!

### Counter-Example: When we might reject

If the proposed value had lower likelihood:
- Current log-likelihood: -10.0
- Proposed log-likelihood: -12.0
- log_acceptance_ratio = -2.0
- acceptance_prob = exp(-2.0) = 0.135

**Result:** We accept with probability 13.5% and reject with probability 86.5%.

## 5.7 Why This Makes Sense Intuitively

The acceptance ratio is essentially comparing:

$$\text{Posterior}(\phi^*) \cdot \phi^* \quad \text{vs.} \quad \text{Posterior}(\phi) \cdot \phi$$

### Think of it this way:

1. **Prior term**: Does $\phi^*$ make more sense given our prior beliefs?
2. **Likelihood term**: Does $\phi^*$ explain the spatial structure better?
3. **Jacobian term**: Are we accounting for the log-transformation correctly?

The MCMC chain will naturally move toward regions of high posterior probability, while occasionally accepting worse proposals to explore the parameter space.

## 5.8 Practical Implications

### Tuning the Acceptance Rate

The proposal standard deviation `phi_proposal_sd` controls the acceptance rate:

- **Too small**: High acceptance rate (>50%), but slow exploration (random walk)
- **Too large**: Low acceptance rate (<20%), many rejections (inefficient)
- **Just right**: 20-40% acceptance rate (optimal for random-walk MH)

### The Acceptance Rate in the Code

The function tracks:
```{r eval=FALSE}
phi_accepted <- logical(n_iter)
```
and returns:
```{r eval=FALSE}
acceptance_rate = mean(phi_accepted[2:n_iter])
```

A good acceptance rate for this type of proposal is typically 20-40%. If it's outside this range, consider adjusting `phi_proposal_sd`.

### Summary of the Acceptance Ratio

The Metropolis-Hastings acceptance ratio for $\phi$:

$$\alpha = \min\left(1, \exp(\log \pi(\phi^*) + \log L(W|\phi^*) + \log(\phi^*) - \log \pi(\phi) - \log L(W|\phi) - \log(\phi))\right)$$

Is a **fairness criterion** that:
1. Favors values with higher posterior density
2. Sometimes accepts lower-density values to maintain exploration
3. Correctly accounts for the log-transformation via the Jacobian
4. Ensures the Markov chain converges to the correct target distribution

# 6. Complete Mathematical Model Summary

## Full Conditional Distributions:

1. **$\beta \mid Y, W, \tau^2$**: 
   $$\mathcal{MVN}\left(\left(\frac{X^TX}{\tau^2} + P_{\beta}\right)^{-1}\left(\frac{X^T(Y-W)}{\tau^2} + P_{\beta}\beta_{\text{prior}}\right), \left(\frac{X^TX}{\tau^2} + P_{\beta}\right)^{-1}\right)$$

2. **$W \mid Y, \beta, \sigma^2, \tau^2, \phi$**: 
   $$\mathcal{MVN}\left(\left(\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}\right)^{-1}\left(\frac{Y-X\beta}{\tau^2}\right), \left(\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}\right)^{-1}\right)$$

3. **$\sigma^2 \mid W, \phi$**: 
   $$\text{Inverse-Gamma}\left(\alpha_{\sigma} + \frac{n}{2}, \beta_{\sigma} + \frac{1}{2}W^T R^{-1} W\right)$$

4. **$\tau^2 \mid Y, \beta, W$**: 
   $$\text{Inverse-Gamma}\left(\alpha_{\tau} + \frac{n}{2}, \beta_{\tau} + \frac{1}{2}(Y-X\beta-W)^T(Y-X\beta-W)\right)$$

5. **$\phi \mid W, \sigma^2$**: Metropolis-Hastings step on $\eta = \log(\phi)$

## The Target Posterior:

$$\pi(\beta, W, \sigma^2, \tau^2, \phi \mid Y) \propto$$
$$[Y|\beta, W, \tau^2] \times [W|\sigma^2, \phi] \times [\beta] \times [\sigma^2] \times [\tau^2] \times [\phi]$$

Where:
$$[Y|\beta, W, \tau^2] \propto (\tau^2)^{-n/2} \exp\left(-\frac{1}{2\tau^2}(Y-X\beta-W)^T(Y-X\beta-W)\right)$$
$$[W|\sigma^2, \phi] \propto (\sigma^2)^{-n/2}|R(\phi)|^{-1/2} \exp\left(-\frac{1}{2\sigma^2}W^T R(\phi)^{-1}W\right)$$

# 7. Key Insights

## The Precision-Addition Principle

Throughout the MCMC algorithm, we see a recurring theme:

$$\text{Posterior Precision} = \text{Data Precision} + \text{Prior Precision}$$

This appears in:
- **$\beta$ update**: $\frac{X^TX}{\tau^2} + P_{\beta}$
- **$W$ update**: $\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}$

## Why This Matters

1. **Numerical stability**: Working with precisions avoids inverting ill-conditioned matrices
2. **Efficiency**: Precision matrices are often sparse in spatial models
3. **Intuition**: Directly shows how data and prior combine

## The Role of Spatial Parameters

| Parameter | Meaning | Effect |
|:-----------|:---------|:--------|
| $\sigma^2$ | Spatial variance | How much spatial variation exists |
| $\tau^2$ | Error variance | How much independent noise exists |
| $\phi$ | Spatial range | How far spatial correlation extends |

# 8. References and Further Reading

1. Banerjee, S., Carlin, B. P., & Gelfand, A. E. (2014). *Hierarchical Modeling and Analysis for Spatial Data*. CRC Press.

2. Cressie, N. A. C. (2015). *Statistics for Spatial Data*. John Wiley & Sons.

3. Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., & Rubin, D. B. (2013). *Bayesian Data Analysis*. CRC Press.

4. Gelfand, A. E., Diggle, P. J., Fuentes, M., & Guttorp, P. (2010). *Handbook of Spatial Statistics*. CRC Press.

5. Robert, C. P., & Casella, G. (2004). *Monte Carlo Statistical Methods*. Springer.

# 9. Exercises

1. **Modify the priors**: Change the prior for $\phi$ and observe how the MCMC output changes
2. **Explore acceptance rates**: Adjust `phi_proposal_sd` and monitor the acceptance rate
3. **Add a new covariance function**: Implement a spherical correlation function
4. **Prediction**: Use the posterior samples to predict at new locations
5. **Convergence diagnostics**: Use the `coda` package to check MCMC convergence
6. **Sensitivity analysis**: Change the prior hyperparameters and see how posterior estimates change
7. **Visualization**: Create trace plots and density plots for all parameters


## Full R code explanations


# 1. The Statistical Model

The model assumes:

$$Y = X\beta + W + \varepsilon$$

Where:
- $Y$: Response variable ($n \times 1$)
- $X$: Design matrix ($n \times p$) of covariates
- $\beta$: Regression coefficients ($p \times 1$)
- $W$: Spatial random effects ($n \times 1$) with $W \sim \mathcal{MVN}(0, \sigma^2 R(\phi))$
- $\varepsilon$: Independent errors with $\varepsilon \sim \mathcal{MVN}(0, \tau^2 I)$

## 1.1 Spatial Correlation Functions

The spatial correlation matrix $R(\phi)$ depends on the distance between locations:

### Exponential Correlation:
$$R(\phi)_{ij} = \exp\left(-\frac{d_{ij}}{\phi}\right)$$

### Matérn Correlation (with $\kappa = 1.5$):
$$R(\phi)_{ij} = \left(1 + \frac{d_{ij}}{\phi}\right) \exp\left(-\frac{d_{ij}}{\phi}\right)$$

Where $d_{ij}$ is the distance between locations $i$ and $j$, and $\phi$ is the spatial range parameter.

# 2. Prior Distributions

The function uses conjugate priors:

## Regression Coefficients ($\beta$)
$$\beta \sim \mathcal{MVN}(\beta_{\text{mean}}, \beta_{\text{var}} \cdot I)$$
Where $\beta_{\text{mean}} = 0$ and $\beta_{\text{var}} = 1000$ (vague prior)

## Spatial Variance ($\sigma^2$)
$$\sigma^2 \sim \text{Inverse-Gamma}(\alpha_{\sigma} = 2, \beta_{\sigma} = 1)$$

## Error Variance ($\tau^2$)
$$\tau^2 \sim \text{Inverse-Gamma}(\alpha_{\tau} = 2, \beta_{\tau} = 1)$$

## Spatial Range ($\phi$)
$$\phi \sim \text{Gamma}(\alpha_{\phi} = 2, \beta_{\phi} = 1)$$

# 3. The `rmvn_precision` Function

```{r}
rmvn_precision <- function(precision, linear_term) {
  U <- chol(precision)
  posterior_mean <- backsolve(U, forwardsolve(t(U), linear_term))
  random_part <- backsolve(U, rnorm(length(linear_term)))
  as.numeric(posterior_mean + random_part)
}
```

## 3.1 Function Purpose and Inputs

This function samples from a multivariate normal distribution using the **precision matrix** (inverse covariance) parameterization.

**Inputs:**
- `precision`: The precision matrix $Q = \Sigma^{-1}$ (positive definite)
- `linear_term`: The vector $b = Q\mu$ (precision times mean)

**Output:**
- A single draw from $\mathcal{N}(Q^{-1}b, Q^{-1})$

## 3.2 Detailed Code Explanation

### Line 1: Cholesky Decomposition
```{r eval=FALSE}
U <- chol(precision)
```

**What it does:** 
- Computes the Cholesky decomposition of the precision matrix
- Returns an upper triangular matrix $U$ such that $Q = U^T U$

**Why use Cholesky?**
- Triangular matrices are easy to solve (forward/backward substitution)
- More numerically stable than directly inverting $Q$
- Exploits the positive definiteness of precision matrices

**Example:**
```{r}
# Simple 2x2 example
Q <- matrix(c(2, 1, 1, 2), nrow = 2)
U <- chol(Q)
print(U)
# U^T %*% U should equal Q
print(t(U) %*% U)
```

### Line 2: Compute Posterior Mean
```{r eval=FALSE}
posterior_mean <- backsolve(U, forwardsolve(t(U), linear_term))
```

**What it does:**
Solves the linear system $Q\mu = b$ using two steps:

**Step 2a: Forward Solve**
```{r eval=FALSE}
forwardsolve(t(U), linear_term)
```
- Solves $U^T z = b$ for $z$
- $U^T$ is lower triangular (since $U$ is upper triangular)
- Uses forward substitution: $z_i = (b_i - \sum_{j=1}^{i-1} U_{ji}z_j) / U_{ii}$

**Step 2b: Backward Solve**
```{r eval=FALSE}
backsolve(U, z)
```
- Solves $U\mu = z$ for $\mu$
- $U$ is upper triangular
- Uses backward substitution: $\mu_i = (z_i - \sum_{j=i+1}^{n} U_{ij}\mu_j) / U_{ii}$

**Why this is efficient:**
- Triangular solve is $O(n^2)$ vs $O(n^3)$ for matrix inversion
- Exploits the structure without explicitly computing $Q^{-1}$

### Line 3: Generate Random Noise
```{r eval=FALSE}
random_part <- backsolve(U, rnorm(length(linear_term)))
```

**What it does:**
1. Generates standard normal draws: $Z \sim \mathcal{N}(0, I)$
2. Solves $U Y = Z$ to get $Y = U^{-1}Z$

**Why this works:**
$$\text{Var}(Y) = \text{Var}(U^{-1}Z) = U^{-1} \text{Var}(Z)(U^{-1})^T = U^{-1}I(U^{-1})^T = (U^T U)^{-1} = Q^{-1}$$

So $Y \sim \mathcal{N}(0, Q^{-1})$

### Line 4: Combine and Return
```{r eval=FALSE}
as.numeric(posterior_mean + random_part)
```

**What it does:**
- Adds the mean to the random component
- `as.numeric()` ensures we return a vector (not a matrix)

**Final result:**
$$X = \mu + Y \sim \mathcal{N}(\mu, Q^{-1})$$

## 3.3 Complete Example

```{r}
# Set up a 2D normal distribution
Q <- matrix(c(4, 1, 1, 3), nrow = 2)  # Precision matrix
b <- c(2, 3)  # Linear term

# Draw a sample
set.seed(123)
sample <- rmvn_precision(Q, b)
print(sample)

# Verify with direct method
Sigma <- solve(Q)
mu <- Sigma %*% b
set.seed(123)
direct_sample <- MASS::mvrnorm(1, mu, Sigma)
print(direct_sample)
# Should be approximately equal
```

# 4. The Main MCMC Function - Detailed Code Walkthrough

## 4.1 Data Preparation and Validation

```{r eval=FALSE}
run_spatial_mcmc <- function(
    data,
    response,
    covariates,
    coords,
    n_iter = 10000,
    burnin = 5000,
    thin = 5,
    cov_family = "exponential",
    priors = list(
      beta_mean = 0,
      beta_var = 1000,
      sigma2_shape = 2,
      sigma2_rate = 1,
      tau2_shape = 2,
      tau2_rate = 1,
      phi_shape = 2,
      phi_rate = 1
    ),
    initial_values = list(
      beta = NULL,
      W = NULL,
      sigma2 = 0.5,
      tau2 = 0.2,
      phi = 1
    ),
    phi_proposal_sd = 0.15,
    progress_every = 100
) {
```

### Extract and Prepare Data
```{r eval=FALSE}
  # Extract response variable as numeric vector
  Y <- as.numeric(data[[response]])
  
  # Create design matrix from covariates
  X <- model.matrix(
    reformulate(covariates, response = NULL),
    data = data
  )
  
  # Ensure coordinates are in matrix format
  coords <- as.matrix(coords)
```

**Explanation:**
- `data[[response]]`: Extracts the response column using the column name
- `as.numeric()`: Ensures the response is a numeric vector
- `model.matrix()`: Creates the design matrix including an intercept (default)
- `reformulate(covariates, response = NULL)`: Creates formula from character vector
- `as.matrix()`: Converts coordinates to matrix (if they're a data frame)

### Basic Parameter Extraction
```{r eval=FALSE}
  n <- length(Y)
  p <- ncol(X)
  
  if (nrow(X) != n || nrow(coords) != n) {
    stop("Y, X and coordinates must have the same number of rows.")
  }
```

**Explanation:**
- `n`: Number of observations
- `p`: Number of regression coefficients (including intercept)
- Validation ensures all inputs are consistent

### Data Validation Checks
```{r eval=FALSE}
  if (!all(is.finite(Y))) {
    stop("The response contains NA, NaN or Inf.")
  }
  
  if (!all(is.finite(X))) {
    stop("The design matrix contains NA, NaN or Inf.")
  }
  
  if (!all(is.finite(coords))) {
    stop("The coordinates contain NA, NaN or Inf.")
  }
```

**Explanation:**
- `is.finite()`: Checks for NA, NaN, and Inf values
- `all()`: Ensures ALL values are finite
- Prevents errors during matrix operations

### Parameter Validation
```{r eval=FALSE}
  if (n_iter < 2 || burnin < 0 || burnin >= n_iter || thin < 1) {
    stop("Require n_iter >= 2, 0 <= burnin < n_iter, and thin >= 1.")
  }
  
  if (!is.finite(phi_proposal_sd) || phi_proposal_sd <= 0) {
    stop("phi_proposal_sd must be positive and finite.")
  }
```

**Explanation:**
- Ensures MCMC settings are valid
- `thin` must be at least 1 (no thinning if thin = 1)
- `phi_proposal_sd` must be positive for a valid proposal distribution

## 4.2 Coordinate Scaling and Distance Matrix

```{r eval=FALSE}
  # Standardising coordinates makes phi dimensionless
  coords_scaled <- scale(coords)
  coord_center <- attr(coords_scaled, "scaled:center")
  coord_scale <- attr(coords_scaled, "scaled:scale")
  distance_matrix <- as.matrix(dist(coords_scaled))
```

**Explanation:**
- `scale()`: Standardizes coordinates to mean 0, variance 1
  - This makes $\phi$ dimensionless and easier to interpret
  - Improves numerical stability
- `attr(coords_scaled, "scaled:center")`: Saves the centering values
- `attr(coords_scaled, "scaled:scale")`: Saves the scaling values
- `dist()`: Computes pairwise Euclidean distances
- `as.matrix()`: Converts to full distance matrix

**Why scale coordinates?**
- Without scaling, $\phi$ would be in the units of the coordinates
- Scaling makes $\phi$ a unitless parameter (easier to set priors)
- Scaling information is stored for future predictions

## 4.3 Correlation Matrix Function

```{r eval=FALSE}
  make_correlation <- function(phi) {
    if (!is.finite(phi) || phi <= 0) {
      stop("phi must be positive and finite.")
    }
    
    if (cov_family == "exponential") {
      R <- exp(-distance_matrix / phi)
    } else if (cov_family == "matern") {
      R <- geoR::matern(
        distance_matrix,
        phi = phi,
        kappa = 1.5
      )
    } else {
      stop("Use cov_family = 'exponential' or 'matern'.")
    }
    
    # A tiny diagonal jitter protects Cholesky decompositions
    R + diag(1e-8, n)
  }
```

**Explanation:**
- `make_correlation()`: Function closure that uses `distance_matrix` and `cov_family`
- `phi` must be positive (spatial range parameter)

**Exponential correlation:**
```{r eval=FALSE}
R <- exp(-distance_matrix / phi)
```
- Each element is the exponential of negative distance/phi
- Correlation decays exponentially with distance

**Matérn correlation:**
```{r eval=FALSE}
R <- geoR::matern(distance_matrix, phi = phi, kappa = 1.5)
```
- Uses the `geoR` package implementation
- `kappa = 1.5` gives a specific smoothness
- More flexible than exponential

**Jitter:**
```{r eval=FALSE}
R + diag(1e-8, n)
```
- Adds tiny amount to diagonal
- Prevents numerical issues with Cholesky decomposition
- Small enough not to affect results (nugget effect is separate)

## 4.4 Initial Values

```{r eval=FALSE}
  # Beta initial values
  beta_current <- if (is.null(initial_values$beta)) {
    as.numeric(qr.solve(X, Y))
  } else {
    as.numeric(initial_values$beta)
  }
  
  if (length(beta_current) != p || !all(is.finite(beta_current))) {
    stop("initial_values$beta must contain ", p, " finite values.")
  }
```

**Explanation:**
- `if (is.null(initial_values$beta))`: If user didn't provide initial values
- `qr.solve(X, Y)`: Computes least squares estimate (OLS)
  - Uses QR decomposition for numerical stability
  - Good starting point for MCMC
- `as.numeric()`: Ensures vector format
- Validation checks length and finite values

**W (spatial effects) initialization:**
```{r eval=FALSE}
  W_current <- if (is.null(initial_values$W)) {
    rep(0, n)
  } else {
    as.numeric(initial_values$W)
  }
  
  if (length(W_current) != n || !all(is.finite(W_current))) {
    stop("initial_values$W must contain ", n, " finite values.")
  }
```

**Explanation:**
- Default to zeros for spatial effects
- Zeros are a reasonable starting point (no spatial effect)

**Variance parameters:**
```{r eval=FALSE}
  sigma2_current <- initial_values$sigma2
  tau2_current <- initial_values$tau2
  phi_current <- initial_values$phi
  
  positive_initials <- c(
    sigma2 = sigma2_current,
    tau2 = tau2_current,
    phi = phi_current
  )
  
  if (!all(is.finite(positive_initials)) || any(positive_initials <= 0)) {
    stop("Initial sigma2, tau2 and phi must be positive and finite.")
  }
```

**Explanation:**
- All variance parameters must be positive
- `any(positive_initials <= 0)`: Checks for non-positive values
- Good initial values: sigma2 = 0.5, tau2 = 0.2, phi = 1

**Prior setup:**
```{r eval=FALSE}
  beta_prior_mean <- rep(priors$beta_mean, p)
  beta_prior_precision <- diag(1 / priors$beta_var, p)
```

**Explanation:**
- `rep(priors$beta_mean, p)`: Creates vector of prior means
- `diag(1 / priors$beta_var, p)`: Creates diagonal precision matrix
  - Larger `beta_var` = smaller precision = weaker prior

## 4.5 Storage Allocation

```{r eval=FALSE}
  # Allocate storage for all parameters
  beta_chain <- matrix(NA_real_, nrow = n_iter, ncol = p)
  W_chain <- matrix(NA_real_, nrow = n_iter, ncol = n)
  sigma2_chain <- numeric(n_iter)
  tau2_chain <- numeric(n_iter)
  phi_chain <- numeric(n_iter)
  log_likelihood_chain <- numeric(n_iter)
  phi_accepted <- logical(n_iter)
```

**Explanation:**
- `matrix(NA_real_, nrow, ncol)`: Pre-allocates matrices of NAs
  - `NA_real_`: Specifically numeric NAs
  - Pre-allocation is faster than growing vectors
- `numeric(n_iter)`: Creates numeric vector of length n_iter
- `logical(n_iter)`: Creates logical vector for acceptance tracking

**Initial values stored:**
```{r eval=FALSE}
  beta_chain[1, ] <- beta_current
  W_chain[1, ] <- W_current
  sigma2_chain[1] <- sigma2_current
  tau2_chain[1] <- tau2_current
  phi_chain[1] <- phi_current
```

**Initial likelihood:**
```{r eval=FALSE}
  R_current <- make_correlation(phi_current)
  Sigma_Y <- sigma2_current * R_current + tau2_current * diag(n)
  
  log_likelihood_chain[1] <- mvtnorm::dmvnorm(
    Y,
    mean = as.numeric(X %*% beta_current),
    sigma = Sigma_Y,
    log = TRUE
  )
```

**Explanation:**
- `Sigma_Y = sigma2*R + tau2*I`: Total covariance of Y
- `dmvnorm()`: Multivariate normal density
  - `log = TRUE`: Returns log-density (more stable)
- Stored for convergence diagnostics

## 4.6 MCMC Iterations - Complete Loop

```{r eval=FALSE}
  for (i in 2:n_iter) {
    if (progress_every > 0 && i %% progress_every == 0) {
      cat("Iteration:", i, "\n")
    }
```

**Explanation:**
- `for (i in 2:n_iter)`: Loop from 2 to n_iter
- Progress indicator every `progress_every` iterations

### 4.6.1 Update $\beta$

```{r eval=FALSE}
    # beta | Y, W, tau2
    beta_precision <-
      crossprod(X) / tau2_current + beta_prior_precision
    
    beta_linear_term <-
      crossprod(X, Y - W_current) / tau2_current +
      beta_prior_precision %*% beta_prior_mean
    
    beta_current <- rmvn_precision(
      beta_precision,
      beta_linear_term
    )
```

**Detailed Explanation:**

**`crossprod(X) / tau2_current`**
- `crossprod(X) = X^T X`: Computes X transpose times X
- `crossprod()` is more efficient than `t(X) %*% X`
- Dividing by `tau2_current` scales by error variance

**Prior contribution:**
- `beta_prior_precision`: Diagonal matrix with 1/beta_var
- `beta_prior_mean`: Vector of prior means

**Linear term:**
- `crossprod(X, Y - W_current)`: $X^T(Y - W)$
- `crossprod(X, Y - W_current) / tau2_current`: Data contribution
- `beta_prior_precision %*% beta_prior_mean`: Prior contribution

**Sampling:**
- `rmvn_precision()`: Draws from the full conditional

### 4.6.2 Update $W$

```{r eval=FALSE}
    # W | Y, beta, sigma2, tau2, phi
    R_current <- make_correlation(phi_current)
    R_inverse <- chol2inv(chol(R_current))
    
    W_precision <-
      diag(1 / tau2_current, n) + R_inverse / sigma2_current
    
    W_linear_term <-
      as.numeric(Y - X %*% beta_current) / tau2_current
    
    W_current <- rmvn_precision(W_precision, W_linear_term)
```

**Detailed Explanation:**

**Step 1: Compute correlation matrix**
```{r eval=FALSE}
R_current <- make_correlation(phi_current)
```
- Builds correlation matrix using current phi

**Step 2: Invert correlation matrix**
```{r eval=FALSE}
R_inverse <- chol2inv(chol(R_current))
```
- `chol(R_current)`: Cholesky decomposition of R
- `chol2inv()`: Computes inverse from Cholesky factor
- More stable than `solve(R_current)`

**Step 3: Build precision matrix**
```{r eval=FALSE}
W_precision <- diag(1 / tau2_current, n) + R_inverse / sigma2_current
```
- `diag(1 / tau2_current, n)`: Data precision (identity times 1/tau2)
- `R_inverse / sigma2_current`: Spatial prior precision

**Step 4: Build linear term**
```{r eval=FALSE}
W_linear_term <- as.numeric(Y - X %*% beta_current) / tau2_current
```
- `Y - X %*% beta_current`: Residuals after removing fixed effects
- Divided by tau2 (data precision)

**Step 5: Sample**
```{r eval=FALSE}
W_current <- rmvn_precision(W_precision, W_linear_term)
```
- Draw from full conditional

### 4.6.3 Update $\sigma^2$

```{r eval=FALSE}
    # sigma2 | W, phi
    sigma2_shape <- priors$sigma2_shape + n / 2
    sigma2_rate <- priors$sigma2_rate +
      0.5 * drop(crossprod(W_current, R_inverse %*% W_current))
    
    sigma2_current <- 1 / rgamma(
      1,
      shape = sigma2_shape,
      rate = sigma2_rate
    )
```

**Detailed Explanation:**

**Shape parameter:**
```{r eval=FALSE}
sigma2_shape <- priors$sigma2_shape + n / 2
```
- Prior shape (2) + n/2

**Rate parameter:**
```{r eval=FALSE}
sigma2_rate <- priors$sigma2_rate +
  0.5 * drop(crossprod(W_current, R_inverse %*% W_current))
```
- `crossprod(W_current, R_inverse %*% W_current)`: $W^T R^{-1} W$
- `drop()`: Removes dimensions from matrix (returns scalar)
- Prior rate (1) + 0.5 * quadratic form

**Sampling:**
```{r eval=FALSE}
sigma2_current <- 1 / rgamma(1, shape = sigma2_shape, rate = sigma2_rate)
```
- `rgamma()`: Samples from Gamma distribution
- `1 / rgamma()`: Converts Gamma to Inverse-Gamma
- Shape = sigma2_shape, Rate = sigma2_rate

### 4.6.4 Update $\tau^2$

```{r eval=FALSE}
    # tau2 | Y, beta, W
    residual <- as.numeric(Y - X %*% beta_current - W_current)
    tau2_shape <- priors$tau2_shape + n / 2
    tau2_rate <- priors$tau2_rate + 0.5 * sum(residual^2)
    
    tau2_current <- 1 / rgamma(
      1,
      shape = tau2_shape,
      rate = tau2_rate
    )
```

**Detailed Explanation:**

**Compute residuals:**
```{r eval=FALSE}
residual <- as.numeric(Y - X %*% beta_current - W_current)
```
- Residuals after removing fixed and spatial effects

**Shape parameter:**
```{r eval=FALSE}
tau2_shape <- priors$tau2_shape + n / 2
```
- Prior shape (2) + n/2

**Rate parameter:**
```{r eval=FALSE}
tau2_rate <- priors$tau2_rate + 0.5 * sum(residual^2)
```
- `sum(residual^2)`: Sum of squared residuals
- Prior rate (1) + 0.5 * SSR

**Sampling:**
```{r eval=FALSE}
tau2_current <- 1 / rgamma(1, shape = tau2_shape, rate = tau2_rate)
```
- Inverse-Gamma sampling via Gamma

### 4.6.5 Update $\phi$ - Metropolis-Hastings

```{r eval=FALSE}
    # phi | W, sigma2: random-walk MH on eta = log(phi)
    eta_current <- log(phi_current)
    eta_proposed <- rnorm(1, eta_current, phi_proposal_sd)
    phi_proposed <- exp(eta_proposed)
```

**Detailed Explanation:**

**Log transformation:**
```{r eval=FALSE}
eta_current <- log(phi_current)
```
- Transform phi to log scale

**Propose new value:**
```{r eval=FALSE}
eta_proposed <- rnorm(1, eta_current, phi_proposal_sd)
```
- `rnorm(1, mean, sd)`: Draw one normal deviate
- Symmetric proposal (random walk)

**Back-transform:**
```{r eval=FALSE}
phi_proposed <- exp(eta_proposed)
```
- Transform back to original scale
- Always positive!

**Compute correlation matrices:**
```{r eval=FALSE}
R_proposed <- make_correlation(phi_proposed)
```

**Compute log-targets:**
```{r eval=FALSE}
    log_target_current <- 
      mvtnorm::dmvnorm(
        W_current,
        mean = rep(0, n),
        sigma = sigma2_current * R_current,
        log = TRUE
      ) +
      dgamma(
        phi_current,
        shape = priors$phi_shape,
        rate = priors$phi_rate,
        log = TRUE
      ) +
      eta_current
```

**Detailed Explanation of Log-Target:**
1. **Likelihood term**: `dmvnorm(W_current, ..., log = TRUE)`
   - Evaluates $\log L(W|\phi)$
   
2. **Prior term**: `dgamma(phi_current, ..., log = TRUE)`
   - Evaluates $\log \pi(\phi)$
   
3. **Jacobian term**: `eta_current = log(phi_current)`
   - Accounts for log-transformation

Same calculation for proposed value:
```{r eval=FALSE}
    log_target_proposed <- 
      mvtnorm::dmvnorm(
        W_current,
        mean = rep(0, n),
        sigma = sigma2_current * R_proposed,
        log = TRUE
      ) +
      dgamma(
        phi_proposed,
        shape = priors$phi_shape,
        rate = priors$phi_rate,
        log = TRUE
      ) +
      eta_proposed
```

**Acceptance decision:**
```{r eval=FALSE}
    log_acceptance_ratio <- log_target_proposed - log_target_current
    
    if (
      is.finite(log_acceptance_ratio) &&
      log(runif(1)) < log_acceptance_ratio
    ) {
      phi_current <- phi_proposed
      R_current <- R_proposed
      phi_accepted[i] <- TRUE
    }
```

**Detailed Explanation:**
- `log_acceptance_ratio`: Difference in log-targets
- `runif(1)`: Generate one uniform random number
- `log(runif(1)) < log_acceptance_ratio`: Accept if log(u) < delta
  - Equivalent to u < exp(delta)
- On accept: update phi and correlation matrix
- `phi_accepted[i] <- TRUE`: Track acceptance

### 4.6.6 Compute and Store Log-Likelihood

```{r eval=FALSE}
    # Marginal likelihood Y | beta, sigma2, tau2, phi
    Sigma_Y <- sigma2_current * R_current + tau2_current * diag(n)
    
    log_likelihood_current <- mvtnorm::dmvnorm(
      Y,
      mean = as.numeric(X %*% beta_current),
      sigma = Sigma_Y,
      log = TRUE
    )
```

**Explanation:**
- Total covariance: $\text{Var}(Y) = \sigma^2 R + \tau^2 I$
- Evaluate multivariate normal density for Y
- Used for model comparison and diagnostic

### 4.6.7 Validate and Store

```{r eval=FALSE}
    current_values <- c(
      beta_current,
      W_current,
      sigma2_current,
      tau2_current,
      phi_current,
      log_likelihood_current
    )
    
    if (!all(is.finite(current_values))) {
      stop("A non-finite value was generated at iteration ", i, ".")
    }
    
    beta_chain[i, ] <- beta_current
    W_chain[i, ] <- W_current
    sigma2_chain[i] <- sigma2_current
    tau2_chain[i] <- tau2_current
    phi_chain[i] <- phi_current
    log_likelihood_chain[i] <- log_likelihood_current
```

**Explanation:**
- Combine all values for validation
- Check for finite values (prevents catastrophic failure)
- Store all parameters in pre-allocated arrays

## 4.7 Post-Processing: Burn-in and Thinning

```{r eval=FALSE}
  # 5.6 Burn-in and thinning
  keep_idx <- seq(burnin + 1, n_iter, by = thin)
  
  mcmc_samples <- cbind(
    beta_chain[keep_idx, , drop = FALSE],
    sigma2 = sigma2_chain[keep_idx],
    tau2 = tau2_chain[keep_idx],
    phi = phi_chain[keep_idx],
    log_lik = log_likelihood_chain[keep_idx]
  )
  
  colnames(mcmc_samples)[seq_len(p)] <- paste0("beta_", 0:(p - 1))
```

**Detailed Explanation:**

**Burn-in removal:**
```{r eval=FALSE}
keep_idx <- seq(burnin + 1, n_iter, by = thin)
```
- `seq(from, to, by)`: Creates sequence of indices
- Starts after burn-in: `burnin + 1`
- Skips iterations based on `thin`

**Combine samples:**
```{r eval=FALSE}
mcmc_samples <- cbind(
  beta_chain[keep_idx, , drop = FALSE],
  sigma2 = sigma2_chain[keep_idx],
  tau2 = tau2_chain[keep_idx],
  phi = phi_chain[keep_idx],
  log_lik = log_likelihood_chain[keep_idx]
)
```
- `cbind()`: Column-bind all parameters
- `drop = FALSE`: Prevents dropping to vector if p=1
- Named columns for easier access

**Name beta columns:**
```{r eval=FALSE}
colnames(mcmc_samples)[seq_len(p)] <- paste0("beta_", 0:(p - 1))
```
- Creates names like "beta_0", "beta_1", etc.
- beta_0 is typically the intercept

## 4.8 Validation and MCMC Object Creation

```{r eval=FALSE}
  if (!all(is.finite(mcmc_samples))) {
    bad_values <- which(!is.finite(mcmc_samples), arr.ind = TRUE)
    print(bad_values)
    stop("Non-finite values were found in retained MCMC samples.")
  }
  
  mcmc_object <- coda::mcmc(
    mcmc_samples,
    start = burnin + 1,
    thin = thin
  )
```

**Explanation:**
- `which(!is.finite(mcmc_samples), arr.ind = TRUE)`: 
  - Finds location of non-finite values
  - `arr.ind = TRUE`: Returns row and column indices
- `coda::mcmc()`: Creates coda MCMC object
  - Compatible with coda diagnostic functions
  - Start time and thinning information stored

## 4.9 Return Results

```{r eval=FALSE}
  list(
    samples = mcmc_object,
    parameters = list(
      beta = beta_chain[keep_idx, , drop = FALSE],
      W = W_chain[keep_idx, , drop = FALSE],
      sigma2 = sigma2_chain[keep_idx],
      tau2 = tau2_chain[keep_idx],
      phi = phi_chain[keep_idx],
      log_likelihood = log_likelihood_chain[keep_idx]
    ),
    acceptance_rate = mean(phi_accepted[2:n_iter]),
    kept_iterations = keep_idx,
    coordinate_scaling = list(
      center = coord_center,
      scale = coord_scale
    ),
    settings = list(
      n_iter = n_iter,
      burnin = burnin,
      thin = thin,
      cov_family = cov_family,
      phi_proposal_sd = phi_proposal_sd
    )
  )
}
```

**Explanation of Return Structure:**

**`samples`**: 
- coda MCMC object for easy diagnostics
- Use with `summary()`, `traceplot()`, `gelman.diag()`, etc.

**`parameters`**:
- Raw parameter matrices for custom analysis
- `W` matrix: Each row is one MCMC draw of spatial effects

**`acceptance_rate`**:
- Proportion of accepted phi proposals
- Target: 20-40%
- Used for tuning `phi_proposal_sd`

**`kept_iterations`**:
- Indices of iterations after burn-in and thinning
- Useful for time series analysis

**`coordinate_scaling`**:
- Saved for making predictions at new locations
- New coordinates should be scaled using same center/scale

**`settings`**:
- Stores all MCMC settings for reproducibility

# 5. Complete Mathematical Model Summary

## Full Conditional Distributions:

1. **$\beta \mid Y, W, \tau^2$**: 
   $$\mathcal{MVN}\left(\left(\frac{X^TX}{\tau^2} + P_{\beta}\right)^{-1}\left(\frac{X^T(Y-W)}{\tau^2} + P_{\beta}\beta_{\text{prior}}\right), \left(\frac{X^TX}{\tau^2} + P_{\beta}\right)^{-1}\right)$$

2. **$W \mid Y, \beta, \sigma^2, \tau^2, \phi$**: 
   $$\mathcal{MVN}\left(\left(\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}\right)^{-1}\left(\frac{Y-X\beta}{\tau^2}\right), \left(\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}\right)^{-1}\right)$$

3. **$\sigma^2 \mid W, \phi$**: 
   $$\text{Inverse-Gamma}\left(\alpha_{\sigma} + \frac{n}{2}, \beta_{\sigma} + \frac{1}{2}W^T R^{-1} W\right)$$

4. **$\tau^2 \mid Y, \beta, W$**: 
   $$\text{Inverse-Gamma}\left(\alpha_{\tau} + \frac{n}{2}, \beta_{\tau} + \frac{1}{2}(Y-X\beta-W)^T(Y-X\beta-W)\right)$$

5. **$\phi \mid W, \sigma^2$**: Metropolis-Hastings step on $\eta = \log(\phi)$

# 6. Key Insights

## The Precision-Addition Principle

Throughout the MCMC algorithm, we see a recurring theme:

$$\text{Posterior Precision} = \text{Data Precision} + \text{Prior Precision}$$

This appears in:
- **$\beta$ update**: $\frac{X^TX}{\tau^2} + P_{\beta}$
- **$W$ update**: $\frac{1}{\tau^2}I + \frac{R^{-1}}{\sigma^2}$

## Why This Matters

1. **Numerical stability**: Working with precisions avoids inverting ill-conditioned matrices
2. **Efficiency**: Precision matrices are often sparse in spatial models
3. **Intuition**: Directly shows how data and prior combine

## The Role of Spatial Parameters

| Parameter | Meaning | Effect |
|-----------|---------|--------|
| $\sigma^2$ | Spatial variance | How much spatial variation exists |
| $\tau^2$ | Error variance

