---
title: "Spatial Analysis of Lithology Data - Idaho National Engineering Laboratory"
format: 
 html:
    css: styles.css
    toc: true
    toc-location: left
    
editor: visual
---

# 1 Introduction

These notes are based on an exercise from the book Hierarchical Modeling and Analysis for Spatial Data. We use the [Idaho National Engineering Laboratory lithology dataset](https://github.com/enwuliu/datasets/blob/main/lithology.dat) to demonstrate how to conduct spatial analysis using Bayesian linear regression models.

First, we implement the models manually. We then analyse the data using existing R packages. We use log(Thickness) as the response variable and Surface Elevation and A–B Elevation as covariates. We fit univariate Gaussian spatial models with a nugget effect using exponential and Matérn covariance functions.

# 2 Data description

```{r, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(mgcv)         # For GAM smoothing
library(viridisLite)  # For color palettes
library(fields)       # For image.plot with legend

# 2. IMPORT AND CLEAN DATA ---------------------------------------------

# Import the lithology dataset

litho <- read.table(
  "https://raw.githubusercontent.com/enwuliu/datasets/refs/heads/main/lithology.dat",
  header = FALSE,
  skip = 1,
  col.names = c(
    "Well_Name",
    "Northing",
    "Easting",
    "Surf_Elevation",
    "Thickness_AB",
    "Elevation_AB",
    "Thickness_BC",
    "Elevation_BC",
    "Thickness_CD",
    "Elevation_CD",
    "Thickness_DE"
  ),
  na.strings = c("NA", "N/A", "", " "),
  fill = TRUE,
  stringsAsFactors = FALSE
)

litho2 <- litho %>%
  transmute(
    Northing = as.numeric(gsub(",", "", Northing)),
    Easting = as.numeric(gsub(",", "", Easting)),
    Surf_Elevation = as.numeric(gsub(",", "", Surf_Elevation)),
    Thickness = as.numeric(gsub(",", "", Thickness_AB)),
    AB_Elevation = as.numeric(gsub(",", "", Elevation_AB))
  ) %>%
  drop_na() %>%
  filter(Thickness > 0) %>%
  mutate(
    log_Thickness = log(Thickness),
    Surf_Elevation_scaled = as.numeric(scale(Surf_Elevation)),
    AB_Elevation_scaled = as.numeric(scale(AB_Elevation))
  )

model_variables <- c(
  "Easting",
  "Northing",
  "log_Thickness",
  "Surf_Elevation_scaled",
  "AB_Elevation_scaled"
)

if (!all(vapply(litho2[model_variables], function(x) all(is.finite(x)), logical(1)))) {
  stop("At least one model variable contains NA, NaN or Inf after cleaning.")
}


head(litho2)
summary(litho2)


```

# 3 Plot covariates and response

```{r}
# 3. GAM-BASED PLOTTING FUNCTION ----------------------------------------

plot_lithology_gam <- function(data, variable, plot_title, 
                               show_points = TRUE, 
                               n_contours = 10,
                               grid_size = 150,
                               color_palette = viridis,
                               k_smooth = 20) {
  
  # Extract the variable to model
  z <- data[[variable]]
  
  # Fit GAM with thin plate regression splines
  # k controls the maximum degrees of freedom (smoothness)
  formula <- as.formula(paste(variable, "~ s(Northing, Easting, k =", k_smooth, ")"))
  gam_model <- gam(formula, data = data)
  
  # Create prediction grid
  northing_range <- seq(min(data$Northing), max(data$Northing), length.out = grid_size)
  easting_range <- seq(min(data$Easting), max(data$Easting), length.out = grid_size)
  
  # Expand grid for prediction
  pred_grid <- expand.grid(
    Northing = northing_range,
    Easting = easting_range
  )
  
  # Predict on the grid
  pred_grid$value <- predict(gam_model, newdata = pred_grid, type = "response")
  
  # Convert to matrix format for image/contour functions
  # Note: image() expects x = columns, y = rows, so we need to reshape
  value_matrix <- matrix(pred_grid$value, 
                         nrow = length(easting_range), 
                         ncol = length(northing_range),
                         byrow = FALSE)
  
  # Calculate proper aspect ratio
  asp_ratio <- diff(range(data$Northing)) / diff(range(data$Easting))
  
  # Determine z range for consistent color scaling
  zlim <- range(pred_grid$value, na.rm = TRUE)
  
  # Create the image plot with legend using fields package
  if (requireNamespace("fields", quietly = TRUE)) {
    # Plot with color legend
    fields::image.plot(
      x = easting_range,
      y = northing_range,
      z = value_matrix,
      col = color_palette(100),
      xlab = "Easting (ft)",
      ylab = "Northing (ft)",
      main = plot_title,
      asp = 1/asp_ratio,
      zlim = zlim,
      legend.width = 1.2,
      legend.mar = 3.5,
      legend.args = list(text = variable, side = 4, line = 2.5, cex = 0.8)
    )
  } else {
    # Fallback: plot without legend
    image(
      x = easting_range,
      y = northing_range,
      z = value_matrix,
      col = color_palette(100),
      xlab = "Easting (ft)",
      ylab = "Northing (ft)",
      main = plot_title,
      asp = 1/asp_ratio,
      zlim = zlim
    )
  }
  
  # Add contour lines
  contour(
    x = easting_range,
    y = northing_range,
    z = value_matrix,
    add = TRUE,
    drawlabels = TRUE,
    col = "black",
    lwd = 0.8,
    nlevels = n_contours,
    labcex = 0.6
  )
  
  # Add observed borehole locations
  if (show_points) {
    points(
      data$Easting,
      data$Northing,
      pch = 16,
      cex = 0.4,
      col = adjustcolor("white", alpha.f = 0.6)
    )
  }
  
  box()
  
  # Return the model and predictions for further analysis
  invisible(list(
    model = gam_model,
    grid = pred_grid,
    matrix = value_matrix,
    x = easting_range,
    y = northing_range,
    zlim = zlim
  ))
}

```

## Plot 1: A-B Thickness

```{r, message=FALSE,warning=FALSE}

result_thickness <- plot_lithology_gam(
  data = litho2,
  variable = "Thickness",
  plot_title = "A-B Thickness",
  n_contours = 10,
  grid_size = 150,
  color_palette = viridis,
  k_smooth = 20
)

# Plot 2: Surface Elevation
result_surface <- plot_lithology_gam(
  data = litho2,
  variable = "Surf_Elevation",
  plot_title = "Surface Elevation",
  n_contours = 10,
  grid_size = 150,
  color_palette = viridis,
  k_smooth = 20
)

# Plot 3: A-B Elevation
result_ab <- plot_lithology_gam(
  data = litho2,
  variable = "AB_Elevation",
  plot_title = "A-B Elevation",
  n_contours = 10,
  grid_size = 150,
  color_palette = viridis,
  k_smooth = 20
)

```

# 4 Exploratory data analysis

## 4.1 variable description

```{r, message=FALSE, warning=FALSE}
hist(
  litho2$Thickness,
  main = "Thickness",
  xlab = "Thickness",
  col = "lightblue",
  border = "white"
)

hist(
  litho2$log_Thickness,
  main = "Log thickness",
  xlab = "Log(Thickness)",
  col = "lightgreen",
  border = "white"
)

plot(
  litho2$Surf_Elevation,
  litho2$log_Thickness,
  xlab = "Surface elevation",
  ylab = "Log(Thickness)",
  main = "Surface elevation vs log thickness",
  pch = 19,
  col = "steelblue"
)

plot(
  litho2$AB_Elevation,
  litho2$log_Thickness,
  xlab = "A-B elevation",
  ylab = "Log(Thickness)",
  main = "A-B elevation vs log thickness",
  pch = 19,
  col = "steelblue"
)

colour_group <- cut(
  litho2$log_Thickness,
  breaks = 10,
  include.lowest = TRUE
)

point_colours <- hcl.colors(10, palette = "YlOrRd")[colour_group]

plot(
  litho2$Easting,
  litho2$Northing,
  pch = 19,
  col = point_colours,
  cex = 0.8 +
    1.8 *
      (litho2$log_Thickness - min(litho2$log_Thickness)) /
      diff(range(litho2$log_Thickness)),
  asp = 1,
  xlab = "Easting",
  ylab = "Northing",
  main = "Spatial distribution of log thickness"
)

boxplot(
  litho2$log_Thickness,
  ylab = "Log(Thickness)",
  main = "Log thickness",
  col = "wheat"
)

```

## 4.1 Empirical variogram of the raw response

```{r, message=FALSE, warning=FALSE}

geo_data <- geoR::as.geodata(
  litho2,
  coords.col = match(c("Easting", "Northing"), names(litho2)),
  data.col = match("log_Thickness", names(litho2)),
  covar.col = match(
    c("Surf_Elevation", "AB_Elevation"),
    names(litho2)
  )
)

maximum_distance <- max(
  dist(litho2[, c("Easting", "Northing")])
) * 0.60

variog_raw <- geoR::variog(
  geo_data,
  max.dist = maximum_distance
)

plot(
  variog_raw,
  main = "Empirical variogram of log thickness"
)
```

## 4.2 NON-SPATIAL LINEAR MODEL

```{r, message=FALSE, warning=FALSE}

lm_model <- lm(
  log_Thickness ~ Surf_Elevation_scaled + AB_Elevation_scaled,
  data = litho2
)

print(summary(lm_model))
```

## 4.3 A residual variogram is more appropriate when covariate trends are present.

```{r, message=FALSE, warning=FALSE}

residual_data <- data.frame(
  Easting = litho2$Easting,
  Northing = litho2$Northing,
  residual = residuals(lm_model)
)

geo_residuals <- geoR::as.geodata(
  residual_data,
  coords.col = 1:2,
  data.col = 3
)

variog_residuals <- geoR::variog(
  geo_residuals,
  max.dist = maximum_distance
)

plot(
  variog_residuals,
  main = "Variogram of non-spatial model residuals"
)

```

# 5 The linear spatial model

The model assumes:

$$Y = X\beta + W + \varepsilon$$

Where:

- $Y$: Response variable ($n \times 1$)
- $X$: Design matrix ($n \times p$) of covariates
- $\beta$: Regression coefficients ($p \times 1$)
- $W$: Spatial random effects ($n \times 1$) with $W \sim \mathcal{MVN}(0, \sigma^2 R(\phi))$
- $\varepsilon$: Independent errors with $\varepsilon \sim \mathcal{MVN}(0, \tau^2 I)$


This is the hierarchy of spatial model



- Level 1: DATA MODEL (Likelihood)

$$
Y(s) = X(s)\boldsymbol{\beta} + W(s) + \varepsilon(s)
$$

$$
\varepsilon(s) \sim N(0, \tau^2)
$$

This is what we observe.

- Level 2: PROCESS MODEL (Prior for $W$)

$$
W(s) \sim \operatorname{GP}\left(0, \sigma^2 R(\phi)\right)
$$

This is the underlying spatial process.

- Level 3: PARAMETER MODEL (Hyperpriors)

$$
\sigma^2 \sim \operatorname{IG}(a_\sigma, b_\sigma)
$$

$$
\phi \sim \operatorname{Gamma}(a_\phi, b_\phi)
$$

This is what we don't know.


## 5.1 Spatial Correlation Functions

The spatial correlation matrix $R(\phi)$ depends on the distance between locations:

### Exponential Correlation:

$$R(\phi)_{ij} = \exp\left(-\frac{d_{ij}}{\phi}\right)$$

### Matérn Correlation (with $\kappa = 1.5$):

$$R(\phi)_{ij} = \left(1 + \frac{d_{ij}}{\phi}\right) \exp\left(-\frac{d_{ij}}{\phi}\right)$$

Where $d_{ij}$ is the distance between locations $i$ and $j$, and $\phi$ is the spatial range parameter.

# 6 Prior Distributions

The function uses conjugate priors:

## Regression Coefficients ($\beta$)

$$\beta \sim \mathcal{MVN}(\beta_{\text{mean}}, \beta_{\text{var}} \cdot I)$$ Where $\beta_{\text{mean}} = 0$ and $\beta_{\text{var}} = 1000$ (vague prior)

## Spatial Variance ($\sigma^2$)

$$\sigma^2 \sim \text{Inverse-Gamma}(\alpha_{\sigma} = 2, \beta_{\sigma} = 1)$$

## Error Variance ($\tau^2$)

$$\tau^2 \sim \text{Inverse-Gamma}(\alpha_{\tau} = 2, \beta_{\tau} = 1)$$

## Spatial Range ($\phi$)

$$\phi \sim \text{Gamma}(\alpha_{\phi} = 2, \beta_{\phi} = 1)$$

# 7 Helper functions

## 7.1 The rmvn_precision\` Function

```{r, eval=FALSE}
rmvn_precision <- function(precision, linear_term) {
# precision: = Q (posterior precision)

#   linear_term:b = coefficient of β in the linear term of log-posterior
#   = XᵀY/σ² + P₀μ₀
#   = evidence from data + evidence from prior
# 
# This b is NOT the regression coefficient β itself,
# but rather the COEFFICIENT that multiplies β in the log-posterior
#   
  
  U <- chol(precision)
  # This gives μ = Q⁻¹b (mean of the coefficients β), for cholesky we use forward and backward two steps to solve this equation for μ
  posterior_mean <- backsolve(U, forwardsolve(t(U), linear_term))
  
 
  # This generates random β ~ N(μ, Q⁻¹)
  random_part <- backsolve(U, rnorm(length(linear_term)))
  
   # Each sample is a draw of β (the regression coefficients!)
  as.numeric(posterior_mean + random_part)
```

This function samples from a multivariate normal distribution parameterized by its **precision matrix** (inverse covariance matrix):

$$X \sim \mathcal{N}(Q^{-1}b, Q^{-1})$$ Note: this function can also be used for any other parameter whose posterior has the precision-form normal distribution

$$\theta \mid -\sim N(Q^{-1}\mathbf{b}, Q^{-1})$$ including spatial random effect $w$ when its full conditional is expressed in this form.

## 7.2 Bayesian Linear Regression Model

For the Bayesian linear regression model, assume the following prior distribution:

$$\boldsymbol{\beta}
\sim
N\left(\boldsymbol{\mu}_0,\mathbf{P}_0^{-1}\right),$$

where $\mathbf{P}_0$ is the prior precision matrix.

The likelihood is:

$$\mathbf{Y}\mid\boldsymbol{\beta}
\sim
N\left(\mathbf{X}\boldsymbol{\beta},\sigma^2\mathbf{I}\right).$$

The posterior distribution is:

$$\boldsymbol{\beta}\mid\mathbf{Y}
\sim
N\left(\mathbf{Q}^{-1}\mathbf{b},\mathbf{Q}^{-1}\right),$$

where

$$\mathbf{Q}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\sigma^2}
+\mathbf{P}_0$$

is the **posterior precision matrix**, and

$$\mathbf{b}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{Y}}{\sigma^2}
+\mathbf{P}_0\boldsymbol{\mu}_0$$

is the **posterior linear term**.

Therefore, the posterior mean and covariance matrix are:

$$E\left(\boldsymbol{\beta}\mid\mathbf{Y}\right)
=\mathbf{Q}^{-1}\mathbf{b},
\qquad
\operatorname{Var}\left(\boldsymbol{\beta}\mid\mathbf{Y}\right)
=\mathbf{Q}^{-1}.$$

## 7.3 Understanding Cholesky factorization

```{r, eval=FALSE}
posterior_mean <- backsolve(U, forwardsolve(t(U), linear_term))
```

This single line of code solves the linear system

$$\mathbf{Q}\boldsymbol{\mu}=\mathbf{b}$$

for $\boldsymbol{\mu}$ using Cholesky factorization, where:

$$\mathbf{Q}=\mathbf{U}^{\mathsf T}\mathbf{U}$$

is the precision matrix,

$$\mathbf{b} = \mathtt{linear\_term}$$

is the posterior linear term, and

$$\boldsymbol{\mu}=\mathtt{posterior\_mean}$$

is the quantity we want to calculate.

### 7.3.1 Mathematical Goal

We want to solve:

$$\mathbf{Q}\boldsymbol{\mu}=\mathbf{b}.$$

Substituting $\mathbf{Q}=\mathbf{U}^{\mathsf T}\mathbf{U}$ gives:

$$\mathbf{U}^{\mathsf T}\mathbf{U}\boldsymbol{\mu}=\mathbf{b}.$$

### 7.3.2 Two-Step Solution Strategy

Instead of explicitly calculating $\mathbf{Q}^{-1}$, we use two triangular solves.

#### Step 1: Solve for $\mathbf{z}$

Solve

$$\mathbf{U}^{\mathsf T}\mathbf{z}=\mathbf{b}$$

for $\mathbf{z}$:

```{r, eval=FALSE}
z <- forwardsolve(t(U), linear_term)
```

Because $\mathbf{U}$ is upper triangular, $\mathbf{U}^{\mathsf T}$ is lower triangular, so `forwardsolve()` is used.

#### Step 2: Solve for $\boldsymbol{\mu}$

Solve

$$\mathbf{U}\boldsymbol{\mu}=\mathbf{z}$$

for $\boldsymbol{\mu}$:

```{r, eval=FALSE}
posterior_mean <- backsolve(U, z)
```

Because $\mathbf{U}$ is upper triangular, `backsolve()` is used.

#### Combined into One Line

The two steps can be combined as:

```{r, eval=FALSE}
posterior_mean <- backsolve(
  U,
  forwardsolve(t(U), linear_term)
)
```

Therefore,

$$\boldsymbol{\mu}
=\mathbf{U}^{-1}\mathbf{U}^{-\mathsf T}\mathbf{b}
=\mathbf{Q}^{-1}\mathbf{b}.$$

This approach is more numerically stable and efficient than explicitly calculating $\mathbf{Q}^{-1}$.

Of course we can solve the equation without Cholesky, but it will be expensive.

```{r, eval=FALSE}
posterior_mean <- solve(Q, b)  # O(n³) - expensive!
```

## 7.4 Another helper function.

This function is quite simple. It is used to calculate the spatial correlation matrix using either the exponential or Matérn correlation ($\kappa=1.5$) function.

```{r, eval=FALSE}
 #  Spatial correlation matrix
  make_correlation <- function(phi) {
    if (!is.finite(phi) || phi <= 0) {
      stop("phi must be positive and finite.")
    }

    if (cov_family == "exponential") {
      R <- exp(-distance_matrix / phi)
    } else if (cov_family == "matern") {
      R <- geoR::matern(
        distance_matrix,
        phi = phi,
        kappa = 1.5
      )
    } else {
      stop("Use cov_family = 'exponential' or 'matern'.")
    }

    # A tiny diagonal jitter protects Cholesky decompositions from rounding
    # problems 
    # It is not the nugget variance tau2-it's purely a computational trick
    R + diag(1e-8, n)
  }

```

# 8 Main MCMC function to run data analysis

```{r, message=FALSE, warning=FALSE}
# MCMC FUNCTION ----------------------------------------------------------

run_spatial_mcmc <- function(
    data,
    response,
    covariates,
    coords,
    n_iter = 10000,
    burnin = 5000,
    thin = 5,
    cov_family = "exponential",
    priors = list(
      beta_mean = 0,
      beta_var = 1000,
      sigma2_shape = 2,
      sigma2_rate = 1,
      tau2_shape = 2,
      tau2_rate = 1,
      phi_shape = 2,
      phi_rate = 1
    ),
    initial_values = list(
      beta = NULL,
      W = NULL,
      sigma2 = 0.5,
      tau2 = 0.2,
      phi = 1
    ),
    phi_proposal_sd = 0.15,  #This is the standard deviation of the proposal distribution used in the Metropolis-Hastings algorithm 
                             #for the spatial range parameter ϕ,It controls how large the steps are when proposing new values of ϕ.
    progress_every = 100     #This controls how often the MCMC algorithm prints progress updates to the console during the simulation.
) {                          #i.e we will see Iteration: 100 Iteration: 200 Iteration: 300, ...

  # Data and design matrix
  
  Y <- as.numeric(data[[response]])   

  X <- model.matrix(
    reformulate(covariates, response = NULL),  # create design matrix with intercept
    data = data
  )

  coords <- as.matrix(coords)

  n <- length(Y)
  p <- ncol(X)

  if (nrow(X) != n || nrow(coords) != n) {
    stop("Y, X and coordinates must have the same number of rows.")
  }

  if (!all(is.finite(Y))) {
    stop("The response contains NA, NaN or Inf.")
  }

  if (!all(is.finite(X))) {
    stop("The design matrix contains NA, NaN or Inf.")
  }

  if (!all(is.finite(coords))) {
    stop("The coordinates contain NA, NaN or Inf.")
  }

  if (n_iter < 2 || burnin < 0 || burnin >= n_iter || thin < 1) {
    stop("Require n_iter >= 2, 0 <= burnin < n_iter, and thin >= 1.")
  }

  if (!is.finite(phi_proposal_sd) || phi_proposal_sd <= 0) {
    stop("phi_proposal_sd must be positive and finite.")
  }

  # Standardising coordinates makes phi dimensionless. The transformation
  # attributes are retained for interpretation and future prediction.
  coords_scaled <- scale(coords)
  coord_center <- attr(coords_scaled, "scaled:center") # This extracts the means used for centering:
  coord_scale <- attr(coords_scaled, "scaled:scale") # This extracts the standard deviations used for scaling:
  
  
  distance_matrix <- as.matrix(dist(coords_scaled)) # We will use the distance matrix for the below function.

  #  Spatial correlation matrix
  make_correlation <- function(phi) {
    if (!is.finite(phi) || phi <= 0) {
      stop("phi must be positive and finite.")
    }

    if (cov_family == "exponential") {
      R <- exp(-distance_matrix / phi)
    } else if (cov_family == "matern") {
      R <- geoR::matern(
        distance_matrix,
        phi = phi,
        kappa = 1.5
      )
    } else {
      stop("Use cov_family = 'exponential' or 'matern'.")
    }

    # A tiny diagonal jitter protects Cholesky decompositions from rounding
    # problems 
    # It is not the nugget variance tau2-it's purely a computational trick
    R + diag(1e-8, n)
  }

  
  # important helper function.
  # Draw N(Q^{-1}b, Q^{-1}) using the Cholesky factor of precision Q.
  rmvn_precision <- function(precision, linear_term) {
    U <- chol(precision)

    posterior_mean <- backsolve(
      U,
      forwardsolve(t(U), linear_term)
    )

    random_part <- backsolve(U, rnorm(length(linear_term)))

    as.numeric(posterior_mean + random_part)
  }
  
  

  # Initial values
  beta_current <- if (is.null(initial_values$beta)) {
    as.numeric(qr.solve(X, Y)) # use QR decomposition to solve linear equation for beta
  } else {
    as.numeric(initial_values$beta)
  }
  

  if (length(beta_current) != p || !all(is.finite(beta_current))) {
    stop("initial_values$beta must contain ", p, " finite values.")
  }

  W_current <- if (is.null(initial_values$W)) { # our default initial value of W is Null,
    rep(0, n)
  } else {
    as.numeric(initial_values$W)
  }

  if (length(W_current) != n || !all(is.finite(W_current))) {
    stop("initial_values$W must contain ", n, " finite values.")
  }

  # use our defualt values
  sigma2_current <- initial_values$sigma2
  tau2_current <- initial_values$tau2
  phi_current <- initial_values$phi
  
  
  

  positive_initials <- c(
    sigma2 = sigma2_current,
    tau2 = tau2_current,
    phi = phi_current
  )

  if (!all(is.finite(positive_initials)) || any(positive_initials <= 0)) {
    stop("Initial sigma2, tau2 and phi must be positive and finite.")
  }

  
  beta_prior_mean <- rep(priors$beta_mean, p)
  beta_prior_precision <- diag(1 / priors$beta_var, p) # check the default value of the fuction.

  # Allocate storage
  beta_chain <- matrix(NA_real_, nrow = n_iter, ncol = p)
  
  W_chain <- matrix(NA_real_, nrow = n_iter, ncol = n)
  
  sigma2_chain <- numeric(n_iter)
  
  tau2_chain <- numeric(n_iter)
  
  phi_chain <- numeric(n_iter)
  
  log_likelihood_chain <- numeric(n_iter)
  
  phi_accepted <- logical(n_iter)  # default as FALSE

  beta_chain[1, ] <- beta_current
  W_chain[1, ] <- W_current
  sigma2_chain[1] <- sigma2_current
  tau2_chain[1] <- tau2_current
  phi_chain[1] <- phi_current

  R_current <- make_correlation(phi_current)  #correlation matrix R, exponential or Matern with parameter phi.
  
  Sigma_Y <- sigma2_current * R_current + tau2_current * diag(n)  ### Total covariance = spatial covariance + nugget covariance, see math explanation
  
    log_likelihood_chain[1] <- mvtnorm::dmvnorm(  #PDF of multivariate normal distribution
    Y,
    mean = as.numeric(X %*% beta_current),
    sigma = Sigma_Y,
    log = TRUE
  )

  cat("\nRunning MCMC for", cov_family, "covariance function...\n")
  cat("Number of observations:", n, "\n")
  cat("Number of regression coefficients:", p, "\n")

  #  MCMC iterations
  for (i in 2:n_iter) {
    
    if (progress_every > 0 && i %% progress_every == 0) {  cat("Iteration:", i, "\n")   }

    # beta | Y, W, tau2
    beta_precision <-
      crossprod(X) / tau2_current + beta_prior_precision

    beta_linear_term <-
      crossprod(X, Y - W_current) / tau2_current +
      beta_prior_precision %*% beta_prior_mean

    beta_current <- rmvn_precision(
      beta_precision,
      beta_linear_term
    )

    # W | Y, beta, sigma2, tau2, phi
    R_current <- make_correlation(phi_current)
    R_inverse <- chol2inv(chol(R_current))

    W_precision <-
      diag(1 / tau2_current, n) + R_inverse / sigma2_current

    W_linear_term <-
      as.numeric(Y - X %*% beta_current) / tau2_current

    W_current <- rmvn_precision(W_precision, W_linear_term)

    # sigma2 | W, phi
    sigma2_shape <- priors$sigma2_shape + n / 2
    sigma2_rate <- priors$sigma2_rate +
      0.5 * drop(crossprod(W_current, R_inverse %*% W_current))

    sigma2_current <- 1 / rgamma(
      1,
      shape = sigma2_shape,
      rate = sigma2_rate
    )

    # tau2 | Y, beta, W
    residual <- as.numeric(Y - X %*% beta_current - W_current)
    tau2_shape <- priors$tau2_shape + n / 2
    tau2_rate <- priors$tau2_rate + 0.5 * sum(residual^2)

    tau2_current <- 1 / rgamma(
      1,
      shape = tau2_shape,
      rate = tau2_rate
    )

    # phi | W, sigma2: random-walk MH on eta = log(phi).
    # The +eta terms are the Jacobians for transforming phi to log(phi).
    # Save the current state so rejection can be shown explicitly.
    phi_old <- phi_current
    R_old <- R_current

    eta_current <- log(phi_current)
    eta_proposed <- rnorm(1, eta_current, phi_proposal_sd)
    phi_proposed <- exp(eta_proposed)
    R_proposed <- make_correlation(phi_proposed)

    log_target_current <- mvtnorm::dmvnorm(
      W_current,
      mean = rep(0, n),
      sigma = sigma2_current * R_current,
      log = TRUE
    ) +
      dgamma(
        phi_current,
        shape = priors$phi_shape,
        rate = priors$phi_rate,
        log = TRUE
      ) +
      eta_current

    log_target_proposed <- mvtnorm::dmvnorm(
      W_current,
      mean = rep(0, n),
      sigma = sigma2_current * R_proposed,
      log = TRUE
    ) +
      dgamma(
        phi_proposed,
        shape = priors$phi_shape,
        rate = priors$phi_rate,
        log = TRUE
      ) +
      eta_proposed

    log_acceptance_ratio <- log_target_proposed - log_target_current

    if (
      is.finite(log_acceptance_ratio) &&
      log(runif(1)) < log_acceptance_ratio
    ) {
      # Accept: move the chain to the proposed state.
      phi_current <- phi_proposed
      R_current <- R_proposed
      phi_accepted[i] <- TRUE
    } else {
      # Reject: repeat the previous state in the Markov chain.
      phi_current <- phi_old
      R_current <- R_old
      phi_accepted[i] <- FALSE
    }

    # Marginal likelihood Y | beta, sigma2, tau2, phi
    Sigma_Y <- sigma2_current * R_current + tau2_current * diag(n)  ## Total covariance = spatial covariance + nugget covariance

    log_likelihood_current <- mvtnorm::dmvnorm(  # PDF of multiviate normal distribution
      Y,
      mean = as.numeric(X %*% beta_current),
      sigma = Sigma_Y,
      log = TRUE
    )

    current_values <- c(
      beta_current,
      W_current,
      sigma2_current,
      tau2_current,
      phi_current,
      log_likelihood_current
    )

    if (!all(is.finite(current_values))) {
      stop("A non-finite value was generated at iteration ", i, ".")
    }

    beta_chain[i, ] <- beta_current
    W_chain[i, ] <- W_current
    sigma2_chain[i] <- sigma2_current
    tau2_chain[i] <- tau2_current
    phi_chain[i] <- phi_current
    log_likelihood_chain[i] <- log_likelihood_current
  }

  # Burn-in and thinning
  keep_idx <- seq(burnin + 1, n_iter, by = thin)

  mcmc_samples <- cbind(
    beta_chain[keep_idx, , drop = FALSE],
    sigma2 = sigma2_chain[keep_idx],
    tau2 = tau2_chain[keep_idx],
    phi = phi_chain[keep_idx],
    log_lik = log_likelihood_chain[keep_idx]
  )

  colnames(mcmc_samples)[seq_len(p)] <- paste0("beta_", 0:(p - 1))

  if (!all(is.finite(mcmc_samples))) {
    bad_values <- which(!is.finite(mcmc_samples), arr.ind = TRUE)
    print(bad_values)
    stop("Non-finite values were found in retained MCMC samples.")
  }

  mcmc_object <- coda::mcmc(
    mcmc_samples,
    start = burnin + 1,
    thin = thin
  )

  list(
    samples = mcmc_object,
    parameters = list(
      beta = beta_chain[keep_idx, , drop = FALSE],
      W = W_chain[keep_idx, , drop = FALSE],
      sigma2 = sigma2_chain[keep_idx],
      tau2 = tau2_chain[keep_idx],
      phi = phi_chain[keep_idx],
      log_likelihood = log_likelihood_chain[keep_idx]
    ),
    acceptance_rate = mean(phi_accepted[2:n_iter]),
    kept_iterations = keep_idx,
    coordinate_scaling = list(
      center = coord_center,
      scale = coord_scale
    ),
    settings = list(
      n_iter = n_iter,
      burnin = burnin,
      thin = thin,
      cov_family = cov_family,
      phi_proposal_sd = phi_proposal_sd
    )
  )
}



```

```{r, eval=FALSE}
Sigma_Y <- sigma2_current * R_current + tau2_current * diag(n)
```

## 8.1 math behind the above R code for covariance of Y

### 8.1.1 The Statistical Model

#### Model Specification

Recall the spatial linear model:

$$\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{W} + \boldsymbol{\varepsilon},$$

where:

- $\mathbf{Y}$: $n \times 1$ response vector
- $\mathbf{X}$: $n \times p$ design matrix
- $\boldsymbol{\beta}$: $p \times 1$ vector of regression coefficients
- $\mathbf{W}$: $n \times 1$ vector of spatial random effects
- $\boldsymbol{\varepsilon}$: $n \times 1$ vector of independent errors (nugget)

#### Variance-Covariance Structure

The total variance of $\mathbf{Y}$, conditional on $\boldsymbol{\beta}$, is:

$$\operatorname{Var}(\mathbf{Y})
= \operatorname{Var}(\mathbf{W})
+ \operatorname{Var}(\boldsymbol{\varepsilon}),$$

because $\mathbf{W}$ and $\boldsymbol{\varepsilon}$ are independent.

#### Components of the Covariance Matrix

##### Component 1: Spatial Variance $\sigma^2\mathbf{R}$

The spatial random effects have covariance:

$$\operatorname{Var}(\mathbf{W})
= \sigma^2\mathbf{R}(\phi),$$

where:

- $\sigma^2$: spatial variance parameter (partial sill)
- $\mathbf{R}(\phi)$: spatial correlation matrix that depends on $\phi$

#### Component 2: Nugget Variance $\tau^2\mathbf{I}$

The independent errors have covariance:

$$\operatorname{Var}(\boldsymbol{\varepsilon})
= \tau^2\mathbf{I},$$

where:

- $\tau^2$: nugget variance
- $\mathbf{I}$: $n \times n$ identity matrix

#### Total Covariance

Therefore,

$$\operatorname{Var}(\mathbf{Y})
= \sigma^2\mathbf{R}(\phi) + \tau^2\mathbf{I}.$$

```{r, eval=FALSE}
log_likelihood_current <- dmvnorm(
    Y,
    mean = as.numeric(X %*% beta_current),
    sigma = Sigma_Y,
    log = TRUE
)
```

The above R code calculate the The multivariate normal log-likelihood :

$$\log L = -\frac{n}{2}\log(2\pi)
          - \frac{1}{2}\log|\Sigma|
          - \frac{1}{2}(Y-\mu)^T \Sigma^{-1}(Y-\mu)$$

Where:

$$\mu = X\beta$$

and

$$\Sigma = \sigma^2 R + \tau^2 I$$

# 9 Mathematics of How $\boldsymbol{\beta}$ Was Updated


# Model specification

The spatial linear model is

$$
\mathbf{Y}
=
\mathbf{X}\boldsymbol{\beta}
+
\mathbf{W}
+
\boldsymbol{\varepsilon},
$$

where:

- $\mathbf{Y}$ is an $n \times 1$ vector of responses;
- $\mathbf{X}$ is an $n \times p$ design matrix;
- $\boldsymbol{\beta}$ is a $p \times 1$ vector of regression coefficients;
- $\mathbf{W}$ is an $n \times 1$ vector of spatial random effects; and
- $\boldsymbol{\varepsilon}$ is an $n \times 1$ vector of independent errors.

## Distributional assumptions

The model assumes

$$
\boldsymbol{\varepsilon}
\sim
N_n\left(\mathbf{0},\tau^2\mathbf{I}_n\right)
$$

and

$$
\mathbf{W}
\sim
N_n\left\{\mathbf{0},\sigma^2\mathbf{R}(\phi)\right\}.
$$

Therefore,

$$
\mathbf{Y}
\mid
\boldsymbol{\beta},\mathbf{W},\tau^2
\sim
N_n\left(
\mathbf{X}\boldsymbol{\beta}+\mathbf{W},
\tau^2\mathbf{I}_n
\right).
$$

## Prior specification

The prior for $\boldsymbol{\beta}$ is multivariate normal:

$$
\boldsymbol{\beta}
\sim
N_p\left(\boldsymbol{\mu}_0,\boldsymbol{\Sigma}_0\right),
$$

where $\boldsymbol{\mu}_0$ is the prior mean vector and
$\boldsymbol{\Sigma}_0$ is the prior covariance matrix.

# Part 1: The likelihood function

## Step 1: Rewrite the model in terms of the errors

From

$$
\mathbf{Y}
=
\mathbf{X}\boldsymbol{\beta}
+
\mathbf{W}
+
\boldsymbol{\varepsilon},
$$

we obtain

$$
\boldsymbol{\varepsilon}
=
\mathbf{Y}
-
\mathbf{X}\boldsymbol{\beta}
-
\mathbf{W}.
$$

## Step 2: Write the density of the errors

Since
$\boldsymbol{\varepsilon}\sim
N_n(\mathbf{0},\tau^2\mathbf{I}_n)$,
its probability density function is

$$
f(\boldsymbol{\varepsilon})
=
\frac{1}
{(2\pi)^{n/2}\lvert\tau^2\mathbf{I}_n\rvert^{1/2}}
\exp\left\{
  -\frac{1}{2}
  \boldsymbol{\varepsilon}^{\mathsf T}
  (\tau^2\mathbf{I}_n)^{-1}
  \boldsymbol{\varepsilon}
\right\}.
$$

## Step 3: Simplify the density

The determinant is

$$
\lvert\tau^2\mathbf{I}_n\rvert^{1/2}
=
(\tau^{2n})^{1/2}
=
\tau^n,
$$

and the inverse is

$$
(\tau^2\mathbf{I}_n)^{-1}
=
\frac{1}{\tau^2}\mathbf{I}_n.
$$

Therefore,

$$
f(\boldsymbol{\varepsilon})
=
\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{
  -\frac{1}{2\tau^2}
  \boldsymbol{\varepsilon}^{\mathsf T}
  \boldsymbol{\varepsilon}
\right\}.
$$

## Step 4: Substitute the expression for the errors

Substituting
$\boldsymbol{\varepsilon}
=\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W}$
gives

$$
f(\boldsymbol{\varepsilon})
=
\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{
  -\frac{1}{2\tau^2}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})^{\mathsf T}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})
\right\}.
$$

## Step 5: Transform from the errors to the response

The likelihood of $\mathbf{Y}$ is

$$
L(\boldsymbol{\beta}\mid\mathbf{Y},\mathbf{W},\tau^2)
=
f_{\mathbf{Y}}
(\mathbf{Y}\mid\boldsymbol{\beta},\mathbf{W},\tau^2).
$$

Because

$$
\boldsymbol{\varepsilon}
=
\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W}
\qquad\text{and}\qquad
\frac{\partial\boldsymbol{\varepsilon}}{\partial\mathbf{Y}}
=
\mathbf{I}_n,
$$

the absolute value of the Jacobian determinant is one. Thus,

$$
f_{\mathbf{Y}}
(\mathbf{Y}\mid\boldsymbol{\beta},\mathbf{W},\tau^2)
=
f_{\boldsymbol{\varepsilon}}
(\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W}).
$$

Therefore,

$$
f(\mathbf{Y}\mid\boldsymbol{\beta},\mathbf{W},\tau^2)
=
\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{
  -\frac{1}{2\tau^2}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})^{\mathsf T}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})
\right\}.
$$

## Step 6: The likelihood as a function of the regression coefficients

Viewed as a function of $\boldsymbol{\beta}$,

$$
L(\boldsymbol{\beta})
=
\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{
  -\frac{1}{2\tau^2}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})^{\mathsf T}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})
\right\}.
$$

## Step 7: The likelihood up to proportionality

When deriving the full conditional distribution, terms that do not depend on
$\boldsymbol{\beta}$ can be omitted:

$$
L(\boldsymbol{\beta})
\propto
\exp\left\{
  -\frac{1}{2\tau^2}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})^{\mathsf T}
  (\mathbf{Y}-\mathbf{X}\boldsymbol{\beta}-\mathbf{W})
\right\}.
$$

The normalising constant is dropped because it does not contain
$\boldsymbol{\beta}$, cancels in ratios, and does not affect the shape of the
conditional distribution.

# Part 2: The full conditional distribution of $\boldsymbol{\beta}$

## Step 1: Expand the likelihood

The quadratic form can be expanded as

$$
\begin{aligned}
&(\mathbf{Y}-\mathbf{W}-\mathbf{X}\boldsymbol{\beta})^{\mathsf T}
(\mathbf{Y}-\mathbf{W}-\mathbf{X}\boldsymbol{\beta})\\
&\quad=
(\mathbf{Y}-\mathbf{W})^{\mathsf T}(\mathbf{Y}-\mathbf{W})
-2\boldsymbol{\beta}^{\mathsf T}\mathbf{X}^{\mathsf T}
(\mathbf{Y}-\mathbf{W})
+\boldsymbol{\beta}^{\mathsf T}\mathbf{X}^{\mathsf T}
\mathbf{X}\boldsymbol{\beta}.
\end{aligned}
$$

After removing terms that are constant with respect to
$\boldsymbol{\beta}$,

$$
L(\boldsymbol{\beta})
\propto
\exp\left\{
  -\frac{1}{2\tau^2}
  \boldsymbol{\beta}^{\mathsf T}\mathbf{X}^{\mathsf T}
  \mathbf{X}\boldsymbol{\beta}
  +\frac{1}{\tau^2}
  \boldsymbol{\beta}^{\mathsf T}\mathbf{X}^{\mathsf T}
  (\mathbf{Y}-\mathbf{W})
\right\}.
$$

## Step 2: Write and expand the prior

The prior density, up to proportionality, is

$$
\pi(\boldsymbol{\beta})
\propto
\exp\left\{
  -\frac{1}{2}
  (\boldsymbol{\beta}-\boldsymbol{\mu}_0)^{\mathsf T}
  \boldsymbol{\Sigma}_0^{-1}
  (\boldsymbol{\beta}-\boldsymbol{\mu}_0)
\right\}.
$$

The quadratic form is

$$
\begin{aligned}
&(\boldsymbol{\beta}-\boldsymbol{\mu}_0)^{\mathsf T}
\boldsymbol{\Sigma}_0^{-1}
(\boldsymbol{\beta}-\boldsymbol{\mu}_0)\\
&\quad=
\boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
\boldsymbol{\beta}
-2\boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
\boldsymbol{\mu}_0
+\boldsymbol{\mu}_0^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
\boldsymbol{\mu}_0.
\end{aligned}
$$

Hence,

$$
\pi(\boldsymbol{\beta})
\propto
\exp\left\{
  -\frac{1}{2}
  \boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
  \boldsymbol{\beta}
  +\boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
  \boldsymbol{\mu}_0
\right\}.
$$

## Step 3: Combine the likelihood and prior

By Bayes' theorem,

$$
\pi(\boldsymbol{\beta}\mid\mathbf{Y},\mathbf{W},\tau^2)
\propto
L(\boldsymbol{\beta})\pi(\boldsymbol{\beta}).
$$

Taking logarithms and omitting additive constants gives

$$
\begin{aligned}
\log\pi(\boldsymbol{\beta}\mid\mathbf{Y},\mathbf{W},\tau^2)
={}&
-\frac{1}{2\tau^2}
\boldsymbol{\beta}^{\mathsf T}\mathbf{X}^{\mathsf T}
\mathbf{X}\boldsymbol{\beta}
+\frac{1}{\tau^2}
\boldsymbol{\beta}^{\mathsf T}\mathbf{X}^{\mathsf T}
(\mathbf{Y}-\mathbf{W})\\
&-\frac{1}{2}
\boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
\boldsymbol{\beta}
+\boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_0^{-1}
\boldsymbol{\mu}_0
+C.
\end{aligned}
$$

## Step 4: Collect terms in $\boldsymbol{\beta}$

Collecting the quadratic and linear terms yields

$$
\begin{aligned}
\log\pi(\boldsymbol{\beta}\mid\mathbf{Y},\mathbf{W},\tau^2)
={}&
-\frac{1}{2}\boldsymbol{\beta}^{\mathsf T}
\left(
  \frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
  +\boldsymbol{\Sigma}_0^{-1}
\right)
\boldsymbol{\beta}\\
&+
\boldsymbol{\beta}^{\mathsf T}
\left[
  \frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
  +\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0
\right]
+C.
\end{aligned}
$$

Define the posterior precision matrix and linear term as

$$
\mathbf{Q}_{\beta}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1}
$$

and

$$
\mathbf{b}_{\beta}
=
\frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0.
$$

Then

$$
\log\pi(\boldsymbol{\beta}\mid\cdot)
=
-\frac{1}{2}\boldsymbol{\beta}^{\mathsf T}
\mathbf{Q}_{\beta}\boldsymbol{\beta}
+\boldsymbol{\beta}^{\mathsf T}\mathbf{b}_{\beta}
+C.
$$

## Step 5: Recognise the multivariate normal form

For

$$
\boldsymbol{\beta}
\sim
N_p(\boldsymbol{\mu}_*,\boldsymbol{\Sigma}_*),
$$

the log-density can be written as

$$
\begin{aligned}
\log\pi(\boldsymbol{\beta})
&=
-\frac{1}{2}
(\boldsymbol{\beta}-\boldsymbol{\mu}_*)^{\mathsf T}
\boldsymbol{\Sigma}_*^{-1}
(\boldsymbol{\beta}-\boldsymbol{\mu}_*)
+C\\
&=
-\frac{1}{2}
\boldsymbol{\beta}^{\mathsf T}\boldsymbol{\Sigma}_*^{-1}
\boldsymbol{\beta}
+\boldsymbol{\beta}^{\mathsf T}
\boldsymbol{\Sigma}_*^{-1}\boldsymbol{\mu}_*
+C.
\end{aligned}
$$

## Step 6: Match coefficients

Matching the quadratic terms gives

$$
\boldsymbol{\Sigma}_*^{-1}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1}
=
\mathbf{Q}_{\beta}.
$$

Therefore,

$$
\boldsymbol{\Sigma}_*
=
\mathbf{Q}_{\beta}^{-1}.
$$

Matching the linear terms gives

$$
\boldsymbol{\Sigma}_*^{-1}\boldsymbol{\mu}_*
=
\frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0
=
\mathbf{b}_{\beta}.
$$

Thus,

$$
\boldsymbol{\mu}_*
=
\boldsymbol{\Sigma}_*
\left[
  \frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
  +\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0
\right]
=
\mathbf{Q}_{\beta}^{-1}\mathbf{b}_{\beta}.
$$

## Step 7: Final result

The full conditional distribution is

$$
\boxed{
\boldsymbol{\beta}
\mid
\mathbf{Y},\mathbf{W},\tau^2
\sim
N_p(\boldsymbol{\mu}_*,\boldsymbol{\Sigma}_*)
}
$$

where

$$
\boxed{
\boldsymbol{\Sigma}_*^{-1}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1}
}
$$

and

$$
\boxed{
\boldsymbol{\mu}_*
=
\boldsymbol{\Sigma}_*
\left[
  \frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
  +\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0
\right].
}
$$

# Part 3: Connection to the R code

## R implementation

The full conditional distribution is implemented in the MCMC function as
follows:

```{r, eval=FALSE}
# Posterior precision matrix
beta_precision <- crossprod(X) / tau2_current +
  beta_prior_precision

# Linear term
beta_linear_term <- crossprod(X, Y - W_current) / tau2_current +
  beta_prior_precision %*% beta_prior_mean

# Draw from the multivariate normal full conditional distribution
beta_current <- rmvn_precision(
  beta_precision,
  beta_linear_term
)
```

The mathematical equivalents are

$$
\mathtt{\beta\_precision}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1},
$$

$$
\mathtt{\beta\_linear\_term}
=
\frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
+\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0,
$$

and

$$
\mathtt{\beta\_current}
\sim
N_p\left(
\mathtt{\beta\_precision}^{-1}*
\mathtt{\beta\_linear\_term}, \text{  } 
\mathtt{\beta\_precision}^{-1}
\right).
$$

## The `rmvn_precision()` function

```{r, eval=FALSE}
rmvn_precision <- function(precision, linear_term) {
  U <- chol(precision)

  posterior_mean <- backsolve(
    U,
    forwardsolve(t(U), linear_term)
  )

  random_part <- backsolve(
    U,
    rnorm(length(linear_term))
  )

  as.numeric(posterior_mean + random_part)
}
```

This function:

1. computes the upper-triangular Cholesky factor
   $\mathbf{Q}=\mathbf{U}^{\mathsf T}\mathbf{U}$;
2. solves for the posterior mean
   $\boldsymbol{\mu}=\mathbf{Q}^{-1}\mathbf{b}$;
3. generates $\mathbf{z}\sim N_p(\mathbf{0},\mathbf{I}_p)$; and
4. returns

$$
\mathbf{x}
=
\boldsymbol{\mu}
+
\mathbf{U}^{-1}\mathbf{z}.
$$

Because
$\operatorname{Var}(\mathbf{U}^{-1}\mathbf{z})
=\mathbf{U}^{-1}\mathbf{U}^{-\mathsf T}
=\mathbf{Q}^{-1}$,
the returned value has distribution

$$
\mathbf{x}
\sim
N_p(\mathbf{Q}^{-1}\mathbf{b},\mathbf{Q}^{-1}).
$$

# Part 4: Special cases

## Case 1: Flat prior

When $\boldsymbol{\Sigma}_0^{-1}\rightarrow\mathbf{0}$,

$$
\boldsymbol{\Sigma}_*^{-1}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
$$

and

$$
\boldsymbol{\mu}_*
=
(\mathbf{X}^{\mathsf T}\mathbf{X})^{-1}
\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W}).
$$

This is the ordinary least-squares estimate of $\boldsymbol{\beta}$,
conditional on $\mathbf{W}$.

## Case 2: Ridge prior

If
$\boldsymbol{\Sigma}_0^{-1}=\lambda\mathbf{I}_p$
and $\boldsymbol{\mu}_0=\mathbf{0}$, then

$$
\boldsymbol{\Sigma}_*^{-1}
=
\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
+\lambda\mathbf{I}_p
$$

and

$$
\boldsymbol{\mu}_*
=
\left(
  \frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}
  +\lambda\mathbf{I}_p
\right)^{-1}
\frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}.
$$

This is the ridge-regression posterior mean, conditional on $\mathbf{W}$.

## Case 3: No spatial effect

When $\mathbf{W}=\mathbf{0}$,

$$
\boldsymbol{\mu}_*
=
\boldsymbol{\Sigma}_*
\left(
  \frac{\mathbf{X}^{\mathsf T}\mathbf{Y}}{\tau^2}
  +\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0
\right),
$$

which is the posterior mean from standard Bayesian linear regression.

# Part 5: Intuitive interpretation

The posterior precision is the sum of the information contributed by the data
and the prior:

$$
\underbrace{\boldsymbol{\Sigma}_*^{-1}}_{\text{posterior precision}}
=
\underbrace{\frac{\mathbf{X}^{\mathsf T}\mathbf{X}}{\tau^2}}
_{\text{data precision}}
+
\underbrace{\boldsymbol{\Sigma}_0^{-1}}
_{\text{prior precision}}.
$$

The posterior information vector is similarly

$$
\underbrace{\mathbf{b}_{\beta}}_{\text{posterior information}}
=
\underbrace{
\frac{\mathbf{X}^{\mathsf T}(\mathbf{Y}-\mathbf{W})}{\tau^2}
}_{\text{data information}}
+
\underbrace{
\boldsymbol{\Sigma}_0^{-1}\boldsymbol{\mu}_0
}_{\text{prior information}}.
$$

Finally,

$$
\boldsymbol{\mu}_*
=
\boldsymbol{\Sigma}_*\mathbf{b}_{\beta}.
$$

The main insights are:

- The data contribution uses $\mathbf{Y}-\mathbf{W}$, the response after removing the current spatial effect.
- The prior contribution pulls the posterior toward   $\boldsymbol{\mu}_0$.
- A smaller nugget variance $\tau^2$ gives more weight to the data.
- As $\boldsymbol{\Sigma}_0^{-1}\rightarrow\mathbf{0}$, the posterior mean approaches the ordinary least-squares solution conditional on $\mathbf{W}$.


# 10 Mathematics of How $W$ Was Updated


## 10.1 Model specification

The spatial linear model is

$$
\mathbf Y = \mathbf X\boldsymbol\beta + \mathbf W + \boldsymbol\varepsilon,
$$

where:

- $\mathbf Y$ is an $n \times 1$ vector of responses;
- $\mathbf X$ is an $n \times p$ design matrix;
- $\boldsymbol\beta$ is a $p \times 1$ vector of regression coefficients;
- $\mathbf W$ is an $n \times 1$ vector of spatial random effects; and
- $\boldsymbol\varepsilon$ is an $n \times 1$ vector of independent errors.

## 10.2 Distributional assumptions

The model assumes

$$
\boldsymbol\varepsilon \sim N_n(\mathbf 0,\tau^2\mathbf I_n)
$$

and

$$
\mathbf W \sim N_n\left\{\mathbf 0,\sigma^2\mathbf R(\phi)\right\}.
$$

Therefore,

$$
\mathbf Y\mid\boldsymbol\beta,\mathbf W,\tau^2
\sim
N_n\left(\mathbf X\boldsymbol\beta+\mathbf W,\tau^2\mathbf I_n\right).
$$

## 10.3 The likelihood function for $\mathbf W$

### 10.3.1 Rewrite the model in terms of errors

From the model, we solve for the errors:

$$
\boldsymbol\varepsilon
=\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W.
$$

### 10.3.2 Write the density of the errors

Since $\boldsymbol\varepsilon\sim N_n(\mathbf 0,\tau^2\mathbf I_n)$, its probability density is

$$
f(\boldsymbol\varepsilon)
=\frac{1}{(2\pi)^{n/2}|\tau^2\mathbf I_n|^{1/2}}
\exp\left\{-\frac12
\boldsymbol\varepsilon^\mathsf T
(\tau^2\mathbf I_n)^{-1}
\boldsymbol\varepsilon\right\}.
$$

### 10.3.3 Simplify the density

The determinant is

$$
|\tau^2\mathbf I_n|^{1/2}
=(\tau^{2n})^{1/2}=\tau^n,
$$

and the inverse is

$$
(\tau^2\mathbf I_n)^{-1}=\frac{1}{\tau^2}\mathbf I_n.
$$

Thus,

$$
f(\boldsymbol\varepsilon)
=\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{-\frac{1}{2\tau^2}
\boldsymbol\varepsilon^\mathsf T\boldsymbol\varepsilon\right\}.
$$

### 10.3.4 Substitute the expression for $\boldsymbol\varepsilon$

Substitution gives

$$
f(\boldsymbol\varepsilon)
=\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{-\frac{1}{2\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)\right\}.
$$

### 10.3.5 Transform from errors to $\mathbf Y$

Because

$$
\boldsymbol\varepsilon=\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W
\quad\text{and}\quad
\frac{\partial\boldsymbol\varepsilon}{\partial\mathbf Y}=\mathbf I_n,
$$

the absolute value of the Jacobian determinant is 1. Therefore,

$$
f(\mathbf Y\mid\boldsymbol\beta,\mathbf W,\tau^2)
=\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{-\frac{1}{2\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)\right\}.
$$

### 10.3.6 View the likelihood as a function of $\mathbf W$

Holding $\boldsymbol\beta$ and $\tau^2$ fixed,

$$
L(\mathbf W)
=\frac{1}{(2\pi)^{n/2}\tau^n}
\exp\left\{-\frac{1}{2\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)\right\}.
$$

Up to a constant that does not depend on $\mathbf W$,

$$
L(\mathbf W)\propto
\exp\left\{-\frac{1}{2\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)\right\}.
$$

The factor $1/\{(2\pi)^{n/2}\tau^n\}$ can be dropped because it does not contain $\mathbf W$, cancels in ratios, and does not affect the shape of the conditional distribution.

### 10.3.7 Expand the likelihood

Let $\mathbf r=\mathbf Y-\mathbf X\boldsymbol\beta$. Then

$$
(\mathbf r-\mathbf W)^\mathsf T(\mathbf r-\mathbf W)
=\mathbf r^\mathsf T\mathbf r
-2\mathbf W^\mathsf T\mathbf r
+\mathbf W^\mathsf T\mathbf W.
$$

After dropping the term $\mathbf r^\mathsf T\mathbf r$, which does not depend on $\mathbf W$,

$$
L(\mathbf W)\propto
\exp\left\{
-\frac{1}{2\tau^2}\mathbf W^\mathsf T\mathbf W
+\frac{1}{\tau^2}\mathbf W^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta)
\right\}.
$$

## 10.4 The full conditional distribution for $\mathbf W$

### 10.4.1 Prior for $\mathbf W$

Ignoring its normalising constant, the prior density is

$$
\pi(\mathbf W)\propto
\exp\left\{-\frac12\mathbf W^\mathsf T
(\sigma^2\mathbf R)^{-1}\mathbf W\right\}.
$$

Because

$$
(\sigma^2\mathbf R)^{-1}=\frac{1}{\sigma^2}\mathbf R^{-1},
$$

we have

$$
\pi(\mathbf W)\propto
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W\right\}.
$$

### 10.4.2 Combine the likelihood and prior

By Bayes' theorem,

$$
\pi(\mathbf W\mid\mathbf Y,\boldsymbol\beta,\sigma^2,\tau^2,\phi)
\propto L(\mathbf W)\pi(\mathbf W).
$$

Taking logarithms and ignoring constants,

$$
\log\pi(\mathbf W\mid\cdot)
=-\frac{1}{2\tau^2}\mathbf W^\mathsf T\mathbf W
+\frac{1}{\tau^2}\mathbf W^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta)
-\frac{1}{2\sigma^2}\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W.
$$

Collecting terms in $\mathbf W$ gives

$$
\log\pi(\mathbf W\mid\cdot)
=-\frac12\mathbf W^\mathsf T
\underbrace{\left(
\frac{1}{\tau^2}\mathbf I_n+
\frac{1}{\sigma^2}\mathbf R^{-1}
\right)}_{\text{posterior precision}}
\mathbf W
+\mathbf W^\mathsf T
\underbrace{\left\{
\frac{1}{\tau^2}(\mathbf Y-\mathbf X\boldsymbol\beta)
\right\}}_{\text{linear term}}.
$$

### 10.4.3 Recognise the multivariate normal form

For $\mathbf W\sim N_n(\boldsymbol\mu_W^*,\boldsymbol\Sigma_W^*)$, the log density is

$$
\log\pi(\mathbf W)
=-\frac12
(\mathbf W-\boldsymbol\mu_W^*)^\mathsf T
(\boldsymbol\Sigma_W^*)^{-1}
(\mathbf W-\boldsymbol\mu_W^*)+\text{constant}.
$$

Expanding it gives

$$
\log\pi(\mathbf W)
=-\frac12\mathbf W^\mathsf T
(\boldsymbol\Sigma_W^*)^{-1}\mathbf W
+\mathbf W^\mathsf T
(\boldsymbol\Sigma_W^*)^{-1}\boldsymbol\mu_W^*
+\text{constant}.
$$

Matching the quadratic terms yields

$$
(\boldsymbol\Sigma_W^*)^{-1}
=\frac{1}{\tau^2}\mathbf I_n
+\frac{1}{\sigma^2}\mathbf R^{-1}.
$$

Matching the linear terms yields

$$
(\boldsymbol\Sigma_W^*)^{-1}\boldsymbol\mu_W^*
=\frac{1}{\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta),
$$

and hence

$$
\boldsymbol\mu_W^*
=\boldsymbol\Sigma_W^*
\left\{\frac{1}{\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta)\right\}.
$$

### 10.4.4 Final result

The full conditional distribution is

$$
\mathbf W\mid\mathbf Y,\boldsymbol\beta,\sigma^2,\tau^2,\phi
\sim N_n(\boldsymbol\mu_W^*,\boldsymbol\Sigma_W^*),
$$

where

$$
(\boldsymbol\Sigma_W^*)^{-1}
=\frac{1}{\tau^2}\mathbf I_n
+\frac{1}{\sigma^2}\mathbf R^{-1}
$$

and

$$
\boldsymbol\mu_W^*
=\boldsymbol\Sigma_W^*
\left\{\frac{1}{\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta)\right\}.
$$

## 10.5 Alternative derivation using the joint distribution

From the model,

$$
\begin{pmatrix}
\mathbf Y\\
\mathbf W
\end{pmatrix}
\sim
N_{2n}\left\{
\begin{pmatrix}
\mathbf X\boldsymbol\beta\\
\mathbf 0
\end{pmatrix},
\begin{pmatrix}
\tau^2\mathbf I_n+\sigma^2\mathbf R & \sigma^2\mathbf R\\
\sigma^2\mathbf R & \sigma^2\mathbf R
\end{pmatrix}
\right\}.
$$

The conditional normal formula gives

$$
\mathbf W\mid\mathbf Y
\sim N_n(\boldsymbol\mu_{W\mid Y},\boldsymbol\Sigma_{W\mid Y}),
$$

where

$$
\boldsymbol\mu_{W\mid Y}
=\sigma^2\mathbf R
(\tau^2\mathbf I_n+\sigma^2\mathbf R)^{-1}
(\mathbf Y-\mathbf X\boldsymbol\beta)
$$

and

$$
\boldsymbol\Sigma_{W\mid Y}
=\sigma^2\mathbf R
-\sigma^2\mathbf R
(\tau^2\mathbf I_n+\sigma^2\mathbf R)^{-1}
\sigma^2\mathbf R.
$$

This is equivalent to the precision-based formulation.

## 10.6 Connection to the R code

The conditional distribution is implemented in the MCMC function as follows:

```{r, eval=FALSE}
# Compute R^{-1} using Cholesky decomposition
R_current <- make_correlation(phi_current)
R_inverse <- chol2inv(chol(R_current))

# Build the precision matrix and linear term
W_precision <- diag(1 / tau2_current, n) +
  R_inverse / sigma2_current

W_linear_term <- as.numeric(Y - X %*% beta_current) /
  tau2_current

# Sample from the multivariate normal distribution
W_current <- rmvn_precision(W_precision, W_linear_term)
```

The mathematical equivalents are

$$
\mathtt{W\_precision}
=\frac{1}{\tau^2}\mathbf I_n
+\frac{1}{\sigma^2}\mathbf R^{-1},
$$

$$
\mathtt{W\_linear\_term}
=\frac{1}{\tau^2}(\mathbf Y-\mathbf X\boldsymbol\beta),
$$

and

$$
\mathtt{W\_current}
\sim N_n\left(
\mathtt{W\_precision}^{-1}\mathtt{W\_linear\_term},
\mathtt{W\_precision}^{-1}
\right).
$$

## 10.7 The `rmvn_precision()` function

```{r, eval=FALSE}
rmvn_precision <- function(precision, linear_term) {
  U <- chol(precision)

  posterior_mean <- backsolve(
    U,
    forwardsolve(t(U), linear_term)
  )

  random_part <- backsolve(
    U,
    rnorm(length(linear_term))
  )

  as.numeric(posterior_mean + random_part)
}
```

This function:

1. computes the Cholesky factor $\mathbf Q=\mathbf U^\mathsf T\mathbf U$;
2. solves for the mean $\boldsymbol\mu=\mathbf Q^{-1}\mathbf b$;
3. generates $\mathbf z\sim N_n(\mathbf 0,\mathbf I_n)$; and
4. returns $\mathbf x=\boldsymbol\mu+\mathbf U^{-1}\mathbf z$.

## 10.8 Computing $\mathbf R^{-1}$ using Cholesky decomposition

```{r, eval=FALSE}
# Step 1: Compute the Cholesky factor of R
U <- chol(R_current)

# Step 2: Compute R^{-1}
R_inverse <- chol2inv(U)
```

The Cholesky decomposition is

$$
\mathbf R=\mathbf U^\mathsf T\mathbf U,
$$

where $\mathbf U$ is upper triangular. Therefore,

$$
\mathbf R^{-1}
=\mathbf U^{-1}\mathbf U^{-\mathsf T}
=\mathbf U^{-1}(\mathbf U^\mathsf T)^{-1}.
$$

The `chol2inv()` function obtains this inverse using triangular-system calculations.

### 10.8.1 Intuitive interpretation of Precision matrix

$$
(\boldsymbol\Sigma_W^*)^{-1}
=\frac{1}{\tau^2}\mathbf I_n
+\frac{1}{\sigma^2}\mathbf R^{-1}.
$$

| Term | Source | Interpretation |
|:---|:---|:---|
| $\mathbf I_n/\tau^2$ | Data (likelihood) | Information from the residuals |
| $\mathbf R^{-1}/\sigma^2$ | Prior (process) | Spatial smoothness constraint |

## 10.9 Mean vector

$$
\boldsymbol\mu_W^*
=\boldsymbol\Sigma_W^*
\left\{\frac{1}{\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta)\right\}.
$$

This is a shrinkage estimator. It pulls $\mathbf W$ toward the residuals $\mathbf Y-\mathbf X\boldsymbol\beta$, while smoothing them using the spatial correlation structure. The amount of shrinkage depends on $\sigma^2$, $\tau^2$, and $\phi$.

## 10.10 Relationship to kriging

The conditional mean of $\mathbf W$ is the best linear predictor:

$$
\widehat{\mathbf W}
=\sigma^2\mathbf R
(\tau^2\mathbf I_n+\sigma^2\mathbf R)^{-1}
(\mathbf Y-\mathbf X\boldsymbol\beta).
$$

This is equivalent to the kriging predictor in geostatistics when the model parameters are known.

## 10.11 Special cases

## No spatial variation: $\sigma^2\to 0$

As the spatial variance approaches zero,

$$
\boldsymbol\mu_W^*\to\mathbf 0,
\qquad
\boldsymbol\Sigma_W^*\to\mathbf 0.
$$

Thus, the spatial random effects shrink toward zero.

## 10.12 Vanishing nugget variance: $\tau^2\to 0$

As the independent-error variance approaches zero,

$$
\boldsymbol\mu_W^*
\to\mathbf Y-\mathbf X\boldsymbol\beta,
\qquad
\boldsymbol\Sigma_W^*\to\mathbf 0.
$$

Thus, $\mathbf W$ is forced toward the residual vector.

## 10.13 Known $\boldsymbol\beta$

When $\mathbf X\boldsymbol\beta$ is fixed,

$$
\boldsymbol\mu_W^*
=\boldsymbol\Sigma_W^*
\left\{\frac{1}{\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta)\right\},
$$

which is the standard spatial prediction expression with a known mean.


## 10.14 Note for update $\beta$ and $W$


Both $\beta$ and $W$ are updated using Gibbs sampling because:

- Their full conditional distributions are multivariate normal

- These distributions have closed forms

- They can be sampled from directly using rmvn_precision()

- Every sample is accepted (no rejection)

- No tuning is required


# 11 Mathematics of How $\sigma^2$ Was Updated

## 11.1 The model and assumptions

### 11.1.1 Process model: prior for $\mathbf W$

$$
\mathbf W\mid\sigma^2,\phi
\sim
N_n\left\{\mathbf 0,\sigma^2\mathbf R(\phi)\right\}.
$$

### 11.1.2 Prior for $\sigma^2$

Assume

$$
\sigma^2\sim\operatorname{IG}(a_\sigma,b_\sigma),
$$

using the shape--scale form of the Inverse-Gamma distribution. Its density is

$$
\pi(\sigma^2)
=
\frac{b_\sigma^{a_\sigma}}{\Gamma(a_\sigma)}
(\sigma^2)^{-a_\sigma-1}
\exp\left(-\frac{b_\sigma}{\sigma^2}\right),
\qquad \sigma^2>0,
$$

where:

- $a_\sigma$ is the shape parameter; and
- $b_\sigma$ is the scale parameter in this Inverse-Gamma parameterisation.

> **Parameterisation note:** In the density above, $b_\sigma$ appears in the numerator of $b_\sigma/\sigma^2$. It is therefore conventionally called the Inverse-Gamma **scale** parameter. Equivalently, if $1/\sigma^2$ is Gamma distributed, $b_\sigma$ is its Gamma rate parameter.

## 11.2 The prior density for $\mathbf W$

### 11.2.1 Full expression

From

$$
\mathbf W\mid\sigma^2,\phi
\sim N_n(\mathbf 0,\sigma^2\mathbf R),
$$

the density is

$$
f(\mathbf W\mid\sigma^2,\phi)
=
\frac{1}{(2\pi)^{n/2}|\sigma^2\mathbf R|^{1/2}}
\exp\left\{
-\frac12\mathbf W^\mathsf T
(\sigma^2\mathbf R)^{-1}\mathbf W
\right\}.
$$

- Simplify the determinant

Because $\mathbf R$ is an $n\times n$ matrix,

$$
|\sigma^2\mathbf R|^{1/2}
=\left\{(\sigma^2)^n|\mathbf R|\right\}^{1/2}
=(\sigma^2)^{n/2}|\mathbf R|^{1/2}
=\sigma^n|\mathbf R|^{1/2}.
$$

- Simplify the inverse

$$
(\sigma^2\mathbf R)^{-1}
=\frac{1}{\sigma^2}\mathbf R^{-1}.
$$
- Substitute

Therefore,

$$
f(\mathbf W\mid\sigma^2,\phi)
=
\frac{1}{(2\pi)^{n/2}(\sigma^2)^{n/2}|\mathbf R|^{1/2}}
\exp\left\{
-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
\right\}.
$$

## 11.3 The prior for $\sigma^2$

The Inverse-Gamma prior is

$$
\pi(\sigma^2)
=
\frac{b_\sigma^{a_\sigma}}{\Gamma(a_\sigma)}
(\sigma^2)^{-a_\sigma-1}
\exp\left(-\frac{b_\sigma}{\sigma^2}\right).
$$

In general, if

$$
X\sim\operatorname{IG}(\alpha,\beta),
$$

then

$$
f(x)
=
\frac{\beta^\alpha}{\Gamma(\alpha)}
x^{-\alpha-1}
\exp\left(-\frac{\beta}{x}\right),
\qquad x>0.
$$

Thus, in the present notation,

$$
\alpha=a_\sigma,
\qquad
\beta=b_\sigma.
$$

## 11.4 Combine the density of $\mathbf W$ and the prior for $\sigma^2$

By Bayes' theorem,

$$
\pi(\sigma^2\mid\mathbf W,\phi)
\propto
f(\mathbf W\mid\sigma^2,\phi)\,
\pi(\sigma^2).
$$

## 11.5 Write the unnormalised posterior

$$
\begin{aligned}
\pi(\sigma^2\mid\mathbf W,\phi)
&\propto
\frac{1}{(2\pi)^{n/2}(\sigma^2)^{n/2}|\mathbf R|^{1/2}}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W\right\}\\
&\quad\times
\frac{b_\sigma^{a_\sigma}}{\Gamma(a_\sigma)}
(\sigma^2)^{-a_\sigma-1}
\exp\left(-\frac{b_\sigma}{\sigma^2}\right).
\end{aligned}
$$

- Drop constants not involving $\sigma^2$

The following terms do not depend on $\sigma^2$:

- $(2\pi)^{-n/2}$;
- $|\mathbf R|^{-1/2}$, conditional on $\phi$; and
- $b_\sigma^{a_\sigma}/\Gamma(a_\sigma)$.

They can be absorbed into the proportionality constant. Hence,

$$
\begin{aligned}
\pi(\sigma^2\mid\mathbf W,\phi)
&\propto
(\sigma^2)^{-n/2}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W\right\}\\
&\quad\times
(\sigma^2)^{-a_\sigma-1}
\exp\left(-\frac{b_\sigma}{\sigma^2}\right).
\end{aligned}
$$

- Combine the powers of $\sigma^2$

$$
(\sigma^2)^{-n/2}(\sigma^2)^{-a_\sigma-1}
=
(\sigma^2)^{-(a_\sigma+n/2)-1}.
$$

- Combine the exponential terms

$$
\begin{aligned}
&\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W\right\}
\exp\left(-\frac{b_\sigma}{\sigma^2}\right)\\
&\qquad=
\exp\left\{-\frac{1}{\sigma^2}
\left(b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
\right)\right\}.
\end{aligned}
$$

- Final unnormalised posterior

Therefore,

$$
\pi(\sigma^2\mid\mathbf W,\phi)
\propto
(\sigma^2)^{-(a_\sigma+n/2)-1}
\exp\left\{-\frac{1}{\sigma^2}
\left(b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
\right)\right\}.
$$

## 11.6 Recognise the Inverse-Gamma distribution

The kernel of an Inverse-Gamma distribution is

$$
x^{-\alpha-1}\exp\left(-\frac{\beta}{x}\right).
$$

Comparing this with the posterior kernel gives

$$
x=\sigma^2,
$$

$$
\alpha=a_\sigma+\frac n2,
$$

and

$$
\beta=b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W.
$$

## 11.7 Final result

The full conditional distribution for $\sigma^2$ is

$$
\boxed{
\sigma^2\mid\mathbf W,\phi
\sim
\operatorname{IG}\left(
a_\sigma+\frac n2,
b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
\right)
}.
$$

The updated shape parameter is

$$
a^*=a_\sigma+\frac n2,
$$

and the updated scale parameter is

$$
b^*=b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W.
$$

## 11.8 Connection to the R code

- Code implementation

```{r, eval=FALSE}
# Compute R^{-1} using Cholesky decomposition
R_current <- make_correlation(phi_current)
R_inverse <- chol2inv(chol(R_current))

# Calculate the quadratic form
Wt_Rinv_W <- drop(
  crossprod(W_current, R_inverse %*% W_current)
)

# Updated shape parameter
sigma2_shape <- priors$sigma2_shape + n / 2

# Updated scale for the Inverse-Gamma distribution
# This is also the rate used for the reciprocal Gamma draw.
sigma2_scale <- priors$sigma2_scale + 0.5 * Wt_Rinv_W

# If Z ~ Gamma(shape, rate), then 1/Z ~ IG(shape, scale = rate)
sigma2_current <- 1 / rgamma(
  1,
  shape = sigma2_shape,
  rate = sigma2_scale
)
```

- Mathematical equivalence

The code calculates

$$
\mathtt{sigma2\_shape}
=a_\sigma+\frac n2,
$$

and

$$
\mathtt{sigma2\_scale}
=b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W.
$$

It then generates

$$
\mathtt{sigma2\_current}
\sim
\operatorname{IG}(mathtt{sigma2\_shape},mathtt{sigma2\_scale}).
$$

## 11.9 Why we use the Inverse-Gamma distribution

- Conjugacy

The Inverse-Gamma prior is conjugate for $\sigma^2$ because:

1. the conditional model for $\mathbf W$ is Normal with covariance $\sigma^2\mathbf R$;
2. the prior for $\sigma^2$ is Inverse-Gamma; and
3. the resulting full conditional distribution is also Inverse-Gamma.

This is analogous to using:

- a Normal prior for $\boldsymbol\beta$ with a Normal likelihood, producing a Normal full conditional; and
- an Inverse-Gamma prior for $\tau^2$ in a Normal error model, producing an Inverse-Gamma full conditional.

- Properties of the Inverse-Gamma distribution

For $X\sim\operatorname{IG}(\alpha,\beta)$ under the shape--scale parameterisation used here:

| Property | Value |
|:---|:---|
| Support | $x>0$ |
| Mean | $\displaystyle \frac{\beta}{\alpha-1}$, for $\alpha>1$ |
| Mode | $\displaystyle \frac{\beta}{\alpha+1}$ |
| Variance | $\displaystyle \frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$, for $\alpha>2$ |

## 11.10 Intuitive interpretation

- Shape parameter

$$
a^*=a_\sigma+\frac n2.
$$

- $a_\sigma$ represents prior information.
- $n/2$ is the contribution from the $n$-dimensional spatial random-effects vector.

- Scale parameter

$$
b^*=b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W.
$$

- $b_\sigma$ is the prior scale parameter.
- $\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W$ is the contribution from the spatial random effects.

- Interpretation of the quadratic form

The term

$$
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
$$

is a squared Mahalanobis distance:

1. $\mathbf R^{-1}$ accounts for the correlation among the spatial random effects.
2. The quadratic form measures how large $\mathbf W$ is relative to the variation implied by $\mathbf R$.
3. A larger quadratic form leads to a larger posterior value of $b^*$ and generally supports greater spatial variance.

## 11.11 Full derivation chain

- Step 1: Model assumption

$$
\mathbf W\mid\sigma^2,\phi
\sim N_n(\mathbf 0,\sigma^2\mathbf R).
$$

- Step 2: Density of $\mathbf W$

$$
f(\mathbf W\mid\sigma^2,\phi)
=
\frac{1}{(2\pi)^{n/2}(\sigma^2)^{n/2}|\mathbf R|^{1/2}}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W\right\}.
$$
- Step 3: Prior for $\sigma^2$

$$
\pi(\sigma^2)
=
\frac{b_\sigma^{a_\sigma}}{\Gamma(a_\sigma)}
(\sigma^2)^{-a_\sigma-1}
\exp\left(-\frac{b_\sigma}{\sigma^2}\right).
$$

- Step 4: Apply Bayes' theorem

$$
\pi(\sigma^2\mid\mathbf W,\phi)
\propto
f(\mathbf W\mid\sigma^2,\phi)\pi(\sigma^2).
$$

- Step 5: Simplify

$$
\pi(\sigma^2\mid\mathbf W,\phi)
\propto
(\sigma^2)^{-(a_\sigma+n/2)-1}
\exp\left\{-\frac{1}{\sigma^2}
\left(b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
\right)\right\}.
$$

- Step 6: Recognise the Inverse-Gamma distribution

$$
\boxed{
\sigma^2\mid\mathbf W,\phi
\sim
\operatorname{IG}\left(
a_\sigma+\frac n2,
b_\sigma+
\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W
\right)
}.
$$

# 12. Mathematics of How $\tau^2$ Was Updated

## 12.1 The model and assumptions

- Data model (likelihood)

$$
\mathbf Y\mid\boldsymbol\beta,\mathbf W,\tau^2
\sim
N_n\left(\mathbf X\boldsymbol\beta+\mathbf W,
\tau^2\mathbf I_n\right).
$$

- Prior for $\tau^2$

Assume

$$
\tau^2\sim\operatorname{IG}(a_\tau,b_\tau),
$$

using the shape--scale form of the Inverse-Gamma distribution. Its density is

$$
\pi(\tau^2)
=
\frac{b_\tau^{a_\tau}}{\Gamma(a_\tau)}
(\tau^2)^{-a_\tau-1}
\exp\left(-\frac{b_\tau}{\tau^2}\right),
\qquad \tau^2>0,
$$

where:

- $a_\tau$ is the shape parameter; and
- $b_\tau$ is the scale parameter in this Inverse-Gamma parameterisation.

> **Parameterisation note:** In this density, $b_\tau$ is conventionally called the Inverse-Gamma scale parameter. When $\tau^2$ is generated as the reciprocal of a Gamma random variable, the same value is used as the Gamma rate.

## 12.2 The likelihood function

- Write the likelihood

From the data model, the likelihood as a function of $\tau^2$ is

$$
L(\tau^2)
=
\frac{1}{(2\pi)^{n/2}|\tau^2\mathbf I_n|^{1/2}}
\exp\left\{
-\frac12
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\tau^2\mathbf I_n)^{-1}
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)
\right\}.
$$

- Simplify the determinant

Because $\mathbf I_n$ is an $n\times n$ identity matrix,

$$
|\tau^2\mathbf I_n|^{1/2}
=\left\{(\tau^2)^n\right\}^{1/2}
=(\tau^2)^{n/2}
=\tau^n.
$$

- Simplify the inverse

$$
(\tau^2\mathbf I_n)^{-1}
=\frac{1}{\tau^2}\mathbf I_n.
$$

- Simplify the quadratic form

Define the residual vector

$$
\mathbf r
=\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W.
$$

Then

$$
\mathbf r^\mathsf T
(\tau^2\mathbf I_n)^{-1}\mathbf r
=\frac{1}{\tau^2}\mathbf r^\mathsf T\mathbf r.
$$

- Final likelihood

Therefore,

$$
L(\tau^2)
=
\frac{1}{(2\pi)^{n/2}(\tau^2)^{n/2}}
\exp\left\{
-\frac{1}{2\tau^2}
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)
\right\}.
$$

Equivalently,

$$
L(\tau^2)
=
\frac{1}{(2\pi)^{n/2}(\tau^2)^{n/2}}
\exp\left(-\frac{\mathbf r^\mathsf T\mathbf r}{2\tau^2}\right).
$$

## 12.3 The prior for $\tau^2$

The Inverse-Gamma prior is

$$
\pi(\tau^2)
=
\frac{b_\tau^{a_\tau}}{\Gamma(a_\tau)}
(\tau^2)^{-a_\tau-1}
\exp\left(-\frac{b_\tau}{\tau^2}\right).
$$

## 12.4 Combine the likelihood and prior

By Bayes' theorem,

$$
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
\propto
L(\tau^2)\pi(\tau^2).
$$

- Write the unnormalised posterior

$$
\begin{aligned}
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
&\propto
\frac{1}{(2\pi)^{n/2}(\tau^2)^{n/2}}
\exp\left(-\frac{\mathbf r^\mathsf T\mathbf r}{2\tau^2}\right)\\
&\quad\times
\frac{b_\tau^{a_\tau}}{\Gamma(a_\tau)}
(\tau^2)^{-a_\tau-1}
\exp\left(-\frac{b_\tau}{\tau^2}\right).
\end{aligned}
$$

- Drop constants not involving $\tau^2$

The terms $(2\pi)^{-n/2}$ and $b_\tau^{a_\tau}/\Gamma(a_\tau)$ do not depend on $\tau^2$. They can be absorbed into the proportionality constant:

$$
\begin{aligned}
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
&\propto
(\tau^2)^{-n/2}
\exp\left(-\frac{\mathbf r^\mathsf T\mathbf r}{2\tau^2}\right)\\
&\quad\times
(\tau^2)^{-a_\tau-1}
\exp\left(-\frac{b_\tau}{\tau^2}\right).
\end{aligned}
$$

- Combine the powers of $\tau^2$

$$
(\tau^2)^{-n/2}(\tau^2)^{-a_\tau-1}
=
(\tau^2)^{-(a_\tau+n/2)-1}.
$$

- Combine the exponential terms

$$
\begin{aligned}
&\exp\left(-\frac{\mathbf r^\mathsf T\mathbf r}{2\tau^2}\right)
\exp\left(-\frac{b_\tau}{\tau^2}\right)\\
&\qquad=
\exp\left\{-\frac{1}{\tau^2}
\left(b_\tau+\frac12\mathbf r^\mathsf T\mathbf r\right)
\right\}.
\end{aligned}
$$

- Final unnormalised posterior

Thus,

$$
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
\propto
(\tau^2)^{-(a_\tau+n/2)-1}
\exp\left\{-\frac{1}{\tau^2}
\left(b_\tau+\frac12\mathbf r^\mathsf T\mathbf r\right)
\right\}.
$$

Substituting $\mathbf r=\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W$ gives

$$
\begin{aligned}
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
&\propto
(\tau^2)^{-(a_\tau+n/2)-1}\\
&\quad\times
\exp\left[
-\frac{1}{\tau^2}
\left\{
b_\tau+
\frac12
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)
\right\}
\right].
\end{aligned}
$$

## 12.5 Recognise the Inverse-Gamma distribution

For

$$
X\sim\operatorname{IG}(\alpha,\beta),
$$

the density kernel is

$$
x^{-\alpha-1}\exp\left(-\frac{\beta}{x}\right).
$$

Matching this kernel with the posterior gives

$$
x=\tau^2,
$$

$$
\alpha=a_\tau+\frac n2,
$$

and

$$
\beta=b_\tau+\frac12\mathbf r^\mathsf T\mathbf r.
$$

## 12.5 Final result

The full conditional distribution for $\tau^2$ is

$$
\boxed{
\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W
\sim
\operatorname{IG}\left[
a_\tau+\frac n2,
b_\tau+
\frac12
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T
(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)
\right]
}.
$$

The updated shape parameter is

$$
a^*=a_\tau+\frac n2,
$$

and the updated scale parameter is

$$
b^*=b_\tau+\frac12\mathbf r^\mathsf T\mathbf r,
$$

where

$$
\mathbf r=\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W.
$$

## 12.6 Comparison with $\sigma^2$

- Similarities

Both $\sigma^2$ and $\tau^2$ have Inverse-Gamma full conditional distributions.

| Parameter | Updated shape | Updated scale |
|:---:|:---|:---|
| $\sigma^2$ | $a_\sigma+n/2$ | $b_\sigma+\frac12\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W$ |
| $\tau^2$ | $a_\tau+n/2$ | $b_\tau+\frac12(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)^\mathsf T(\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W)$ |

- Differences

| Aspect | $\sigma^2$ | $\tau^2$ |
|:---|:---|:---|
| Source | Spatial process | Independent error |
| Quadratic form | $\mathbf W^\mathsf T\mathbf R^{-1}\mathbf W$ | $\mathbf r^\mathsf T\mathbf r$ |
| Correlation | Includes spatial correlation through $\mathbf R$ | Assumes independent errors through $\mathbf I_n$ |
| Interpretation | Spatial variance | Nugget variance |

## 12.7 Connection to the R code

- Code implementation

```{r, eval=FALSE}
# Compute residuals
residual <- as.numeric(
  Y - X %*% beta_current - W_current
)

# Updated shape parameter
tau2_shape <- priors$tau2_shape + n / 2

# Updated Inverse-Gamma scale
# This is also the rate for the reciprocal Gamma draw.
tau2_scale <- priors$tau2_scale +
  0.5 * sum(residual^2)

# Sample from the Inverse-Gamma distribution
tau2_current <- 1 / rgamma(
  1,
  shape = tau2_shape,
  rate = tau2_scale
)
```

- Mathematical equivalence

The code calculates

$$
\mathtt{tau2\_shape}
=a_\tau+\frac n2,
$$

and

$$
\mathtt{tau2\_scale}
=b_\tau+\frac12\mathbf r^\mathsf T\mathbf r.
$$

It then generates

$$
\mathtt{tau2\_current}
\sim
\operatorname{IG}(mathtt{tau2\_shape},mathtt{tau2\_scale}).
$$

## 12.8 Step-by-step explanation of the code

- Line 1: Compute the residuals

```{r, eval=FALSE}
residual <- as.numeric(
  Y - X %*% beta_current - W_current
)
```

Mathematically,

$$
\mathbf r=\mathbf Y-\mathbf X\boldsymbol\beta-\mathbf W.
$$

This calculation:

- removes the mean trend $\mathbf X\boldsymbol\beta$;
- removes the spatial random effects $\mathbf W$; and
- leaves the independent errors $\boldsymbol\varepsilon$.

- Line 2: Update the shape parameter

```{r, eval=FALSE}
tau2_shape <- priors$tau2_shape + n / 2
```

Mathematically,

$$
a^*=a_\tau+\frac n2.
$$

The $n/2$ term arises from the factor $(\tau^2)^{-n/2}$ in the Normal likelihood.

- Line 3: Update the scale parameter

```{r, eval=FALSE}
tau2_scale <- priors$tau2_scale +
  0.5 * sum(residual^2)
```

Mathematically,

$$
b^*=b_\tau+\frac12\mathbf r^\mathsf T\mathbf r.
$$

Here:

- `sum(residual^2)` equals $\mathbf r^\mathsf T\mathbf r$;
- the sum of squared residuals is multiplied by $1/2$; and
- the result is added to the prior scale parameter.
- Line 4: Sample $\tau^2$

```{r, eval=FALSE}
tau2_current <- 1 / rgamma(
  1,
  shape = tau2_shape,
  rate = tau2_scale
)
```

If

$$
Z\sim\operatorname{Gamma}(a^*,\text{rate}=b^*),
$$

then

$$
\frac{1}{Z}\sim\operatorname{IG}(a^*,\text{scale}=b^*).
$$

Therefore, `1 / rgamma()` generates the required Inverse-Gamma draw for $\tau^2$.

## 12.9 Full derivation chain

- Step 1: Model assumption

$$
\mathbf Y\mid\boldsymbol\beta,\mathbf W,\tau^2
\sim
N_n(\mathbf X\boldsymbol\beta+\mathbf W,\tau^2\mathbf I_n).
$$

- Step 2: Likelihood for $\tau^2$

$$
L(\tau^2)
\propto
(\tau^2)^{-n/2}
\exp\left(-\frac{\mathbf r^\mathsf T\mathbf r}{2\tau^2}\right).
$$

- Step 3: Prior for $\tau^2$

$$
\pi(\tau^2)
\propto
(\tau^2)^{-a_\tau-1}
\exp\left(-\frac{b_\tau}{\tau^2}\right).
$$

- Step 4: Apply Bayes' theorem

$$
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
\propto
L(\tau^2)\pi(\tau^2).
$$

- Step 5: Simplify

$$
\pi(\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W)
\propto
(\tau^2)^{-(a_\tau+n/2)-1}
\exp\left\{-\frac{1}{\tau^2}
\left(b_\tau+\frac12\mathbf r^\mathsf T\mathbf r\right)
\right\}.
$$
- Step 6: Recognise the Inverse-Gamma distribution

$$
\boxed{
\tau^2\mid\mathbf Y,\boldsymbol\beta,\mathbf W
\sim
\operatorname{IG}\left(
a_\tau+\frac n2,
b_\tau+\frac12\mathbf r^\mathsf T\mathbf r
\right)
}.
$$
#  13 Mathematics as How $\phi$ was updated

## 13.1 The model and assumptions

### 13.1.1 Process model: prior for $\mathbf W$

$$
\mathbf W\mid\sigma^2,\phi
\sim
N_n\left\{\mathbf 0,\sigma^2\mathbf R(\phi)\right\}.
$$

### 13.1.2 Prior for $\phi$

Assume

$$
\phi\sim\operatorname{Gamma}(a_\phi,b_\phi),
$$

where $a_\phi$ is the shape and $b_\phi$ is the rate. The prior density is

$$
\pi(\phi)
=\frac{b_\phi^{a_\phi}}{\Gamma(a_\phi)}
\phi^{a_\phi-1}\exp(-b_\phi\phi),
\qquad \phi>0.
$$

### 13.1.3 The prior density for $\mathbf W$

From the process model,

$$
f(\mathbf W\mid\sigma^2,\phi)
=\frac{1}{(2\pi)^{n/2}|\sigma^2\mathbf R(\phi)|^{1/2}}
\exp\left\{-\frac12\mathbf W^\mathsf T
[\sigma^2\mathbf R(\phi)]^{-1}\mathbf W\right\}.
$$

- Simplify the determinant

$$
|\sigma^2\mathbf R(\phi)|^{1/2}
=\left\{(\sigma^2)^n|\mathbf R(\phi)|\right\}^{1/2}
=(\sigma^2)^{n/2}|\mathbf R(\phi)|^{1/2}.
$$

- Simplify the inverse

$$
[\sigma^2\mathbf R(\phi)]^{-1}
=\frac{1}{\sigma^2}\mathbf R(\phi)^{-1}.
$$

- Substitute

Therefore,

$$
f(\mathbf W\mid\sigma^2,\phi)
=\frac{1}{(2\pi)^{n/2}(\sigma^2)^{n/2}|\mathbf R(\phi)|^{1/2}}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W\right\}.
$$

### 13.1.4 The prior for $\phi$

The Gamma prior is

$$
\pi(\phi)
=\frac{b_\phi^{a_\phi}}{\Gamma(a_\phi)}
\phi^{a_\phi-1}\exp(-b_\phi\phi).
$$

### 13.1.5 Apply Bayes' theorem

By Bayes' theorem,

$$
\pi(\phi\mid\mathbf W,\sigma^2)
\propto
f(\mathbf W\mid\sigma^2,\phi)\pi(\phi).
$$

- Write the unnormalised posterior

$$
\begin{aligned}
\pi(\phi\mid\mathbf W,\sigma^2)
&\propto
\frac{1}{(2\pi)^{n/2}(\sigma^2)^{n/2}|\mathbf R(\phi)|^{1/2}}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W\right\}\\
&\quad\times
\frac{b_\phi^{a_\phi}}{\Gamma(a_\phi)}
\phi^{a_\phi-1}\exp(-b_\phi\phi).
\end{aligned}
$$

- Drop constants not involving $\phi$

Conditional on $\mathbf W$ and $\sigma^2$, the following terms do not depend on $\phi$:

- $(2\pi)^{-n/2}$;
- $(\sigma^2)^{-n/2}$; and
- $b_\phi^{a_\phi}/\Gamma(a_\phi)$.

After dropping them,

$$
\begin{aligned}
\pi(\phi\mid\mathbf W,\sigma^2)
&\propto
|\mathbf R(\phi)|^{-1/2}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W\right\}\\
&\quad\times
\phi^{a_\phi-1}\exp(-b_\phi\phi).
\end{aligned}
$$

- Combine the exponential terms

$$
\pi(\phi\mid\mathbf W,\sigma^2)
\propto
|\mathbf R(\phi)|^{-1/2}
\phi^{a_\phi-1}
\exp\left\{-b_\phi\phi
-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W
\right\}.
$$

- Final unnormalised posterior

The full conditional kernel is

$$
\boxed{
\pi(\phi\mid\mathbf W,\sigma^2)
\propto
|\mathbf R(\phi)|^{-1/2}
\phi^{a_\phi-1}
\exp\left\{-b_\phi\phi
-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W
\right\}
}.
$$

Equivalently,

$$
\pi(\phi\mid\mathbf W,\sigma^2)
\propto
|\sigma^2\mathbf R(\phi)|^{-1/2}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W\right\}
\pi(\phi).
$$

### 13.1.6 Why this has no standard closed form

$\phi$ appears nonlinearly in several parts of the conditional density.

| Term | How $\phi$ appears | Why it is problematic |
|:---|:---|:---|
| $|\mathbf R(\phi)|^{-1/2}$ | Inside the determinant of a correlation matrix | The determinant is a nonlinear function of $\phi$ |
| $\mathbf R(\phi)^{-1}$ | Inside a matrix inverse | Every element of the inverse generally changes with $\phi$ |
| $\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W$ | Inside a quadratic form | It is a nonlinear function of $\phi$ |

### 13.1.7 Determinant term

If $\lambda_i(\phi)$ is the $i$th eigenvalue of $\mathbf R(\phi)$, then

$$
|\mathbf R(\phi)|
=\prod_{i=1}^n\lambda_i(\phi).
$$

Every eigenvalue can depend on $\phi$ in a complicated way.

### 13.1.8 Inverse term

Using an eigendecomposition,

$$
\mathbf R(\phi)^{-1}
=\mathbf U(\phi)\boldsymbol\Lambda(\phi)^{-1}
\mathbf U(\phi)^\mathsf T,
$$

where both the eigenvectors and eigenvalues can depend on $\phi$.

### 13.1.9 Quadratic form

The quadratic form can be written as

$$
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W
=\sum_{i=1}^n
\frac{[\mathbf u_i(\phi)^\mathsf T\mathbf W]^2}
{\lambda_i(\phi)}.
$$

Thus, it is a sum of terms whose numerators and denominators may both depend on $\phi$.

## 13.2 Comparison with the other parameters

| Parameter | Full conditional | Sampling method | Reason |
|:---:|:---|:---|:---|
| $\boldsymbol\beta$ | Multivariate Normal | Gibbs | Normal conjugacy |
| $\mathbf W$ | Multivariate Normal | Gibbs | Normal conjugacy |
| $\sigma^2$ | Inverse-Gamma | Gibbs | Inverse-Gamma conjugacy |
| $\tau^2$ | Inverse-Gamma | Gibbs | Inverse-Gamma conjugacy |
| $\phi$ | Nonstandard | Metropolis--Hastings | No standard conjugate form |

For $\phi$,

$$
f(\mathbf W\mid\sigma^2,\phi)
\propto
|\mathbf R(\phi)|^{-1/2}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W\right\}.
$$

Multiplying this expression by a Gamma prior does not produce a standard named distribution.

## 13.3 The Metropolis--Hastings solution

Metropolis--Hastings only requires the unnormalised target density. Define

$$
h(\phi)
=|\sigma^2\mathbf R(\phi)|^{-1/2}
\exp\left\{-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W\right\}
\pi(\phi).
$$

Then

$$
\pi(\phi\mid\mathbf W,\sigma^2)\propto h(\phi).
$$

- Metropolis--Hastings algorithm

```{r, eval=FALSE}
# Step 1: Transform to the unconstrained log scale
eta_current <- log(phi_current)

# Step 2: Propose a new value
eta_proposed <- rnorm(
  1,
  mean = eta_current,
  sd = phi_proposal_sd
)

# Step 3: Transform back
phi_proposed <- exp(eta_proposed)

# Step 4: Compute log target densities on the eta scale
log_target_current <-
  log_conditional(W, sigma2, phi_current) + eta_current

log_target_proposed <-
  log_conditional(W, sigma2, phi_proposed) + eta_proposed

# Step 5: Compute the log acceptance ratio
log_acceptance_ratio <-
  log_target_proposed - log_target_current

# Step 6: Accept or reject
if (log(runif(1)) < log_acceptance_ratio) {
  phi_current <- phi_proposed
}
```

- R code implementment

```{r, eval=FALSE}
# Random-walk MH on eta = log(phi)
eta_current <- log(phi_current)
eta_proposed <- rnorm(
  1,
  mean = eta_current,
  sd = phi_proposal_sd
)

phi_proposed <- exp(eta_proposed)
R_proposed <- make_correlation(phi_proposed)

# Log target for the current value
log_target_current <- dmvnorm(
  W_current,
  mean = rep(0, n),
  sigma = sigma2_current * R_current,
  log = TRUE
) + dgamma(
  phi_current,
  shape = priors$phi_shape,
  rate = priors$phi_rate,
  log = TRUE
) + eta_current

# Log target for the proposed value
log_target_proposed <- dmvnorm(
  W_current,
  mean = rep(0, n),
  sigma = sigma2_current * R_proposed,
  log = TRUE
) + dgamma(
  phi_proposed,
  shape = priors$phi_shape,
  rate = priors$phi_rate,
  log = TRUE
) + eta_proposed

# Accept or reject
log_acceptance_ratio <-
  log_target_proposed - log_target_current

if (
  is.finite(log_acceptance_ratio) &&
  log(runif(1)) < log_acceptance_ratio
) {
  phi_current <- phi_proposed
  R_current <- R_proposed
  phi_accepted[i] <- TRUE
}
```

## 13.4 Mathematical equivalence

The multivariate Normal log density contributes

$$
-\frac12\log|\sigma^2\mathbf R(\phi)|
-\frac{1}{2\sigma^2}
\mathbf W^\mathsf T\mathbf R(\phi)^{-1}\mathbf W
+\text{constant}.
$$

The Gamma log prior contributes

$$
(a_\phi-1)\log\phi-b_\phi\phi+\text{constant}.
$$

The term $\eta=\log\phi$ is the log-Jacobian adjustment for transforming the target density from $\phi$ to $\eta$.

## 13.5 A complete example

```{r, eval=FALSE}
# Setup
set.seed(123)
n <- 20
coords <- matrix(runif(2 * n), ncol = 2)
dist_mat <- as.matrix(dist(coords))

# True parameters
sigma2_true <- 0.8
phi_true <- 1.5

# Generate W
R <- exp(-dist_mat / phi_true)
L <- chol(sigma2_true * R)
W <- as.numeric(t(L) %*% rnorm(n))

# Prior parameters
priors <- list(phi_shape = 2, phi_rate = 1)

# Compute the unnormalised log conditional density
log_conditional <- function(phi) {
  R <- exp(-dist_mat / phi)
  R_inv <- chol2inv(chol(R))

  log_det <- -0.5 * as.numeric(
    determinant(R, logarithm = TRUE)$modulus
  )

  quad_form <- -0.5 * drop(
    crossprod(W, R_inv %*% W)
  ) / sigma2_true

  log_prior <- dgamma(
    phi,
    shape = priors$phi_shape,
    rate = priors$phi_rate,
    log = TRUE
  )

  log_det + quad_form + log_prior
}

# Evaluate the density on a grid
phi_grid <- seq(0.1, 5, length.out = 100)
log_density <- sapply(phi_grid, log_conditional)
density <- exp(log_density - max(log_density))

# Plot
plot(
  phi_grid,
  density,
  type = "l",
  lwd = 2,
  main = expression("Unnormalised posterior of " * phi),
  xlab = expression(phi),
  ylab = "Relative density"
)

abline(v = phi_true, col = "red", lwd = 2, lty = 2)

legend(
  "topright",
  legend = expression("True " * phi),
  col = "red",
  lty = 2,
  lwd = 2
)
```

## 13.5 Why transform to the log scale?

- Problem with sampling $\phi$ directly

- $\phi$ must be positive: $\phi>0$.
- A Normal random-walk proposal on the original scale can generate negative values.
- Negative proposed values are outside the support of $\phi$.

- Log transformation

Define

$$
\eta=\log\phi.
$$

Then:

- $\eta$ is unconstrained, with $\eta\in(-\infty,\infty)$;
- a Normal proposal for $\eta$ always produces a positive $\phi^*=\exp(\eta^*)$; and
- the transformed target includes a Jacobian adjustment.

## 13.6 Jacobian adjustment

Since

$$
\phi=e^\eta,
\qquad
\frac{d\phi}{d\eta}=e^\eta,
$$

the transformed density is

$$
\pi_\eta(\eta\mid\mathbf W,\sigma^2)
=\pi_\phi(e^\eta\mid\mathbf W,\sigma^2)e^\eta.
$$

On the log scale,

$$
\log\pi_\eta(\eta\mid\cdot)
=\log\pi_\phi(e^\eta\mid\cdot)+\eta.
$$

Therefore, the code includes `+ eta_current` and `+ eta_proposed`.

## 13.7 The acceptance ratio

The Metropolis--Hastings acceptance probability on the transformed scale is

$$
\alpha
=\min\left\{
1,
\frac{\pi_\eta(\eta^*\mid\mathbf W,\sigma^2)
q(\eta\mid\eta^*)}
{\pi_\eta(\eta\mid\mathbf W,\sigma^2)
q(\eta^*\mid\eta)}
\right\}.
$$

The Normal random-walk proposal is symmetric:

$$
q(\eta^*\mid\eta)=q(\eta\mid\eta^*).
$$

Consequently,

$$
\alpha
=\min\left\{
1,
\frac{\pi_\phi(\phi^*\mid\mathbf W,\sigma^2)\phi^*}
{\pi_\phi(\phi\mid\mathbf W,\sigma^2)\phi}
\right\}.
$$

Equivalently, the log acceptance ratio is

$$
\log A
=\log\pi_\phi(\phi^*\mid\cdot)
-\log\pi_\phi(\phi\mid\cdot)
+\eta^*-\eta.
$$

The proposed value is accepted when

$$
\log U<\log A,
\qquad U\sim\operatorname{Uniform}(0,1).
$$

## 13.8 Summary

| Aspect | $\boldsymbol\beta$, $\mathbf W$, $\sigma^2$, $\tau^2$ | $\phi$ |
|:---|:---|:---|
| Full conditional | Standard closed form | Nonstandard form |
| Distribution | Normal or Inverse-Gamma | No standard named distribution |
| Sampling | Gibbs sampling | Metropolis--Hastings |
| Reason | Conjugacy | No standard conjugate prior |
| Acceptance | Direct draw | Proposed draw may be rejected |
| Tuning | None for direct Gibbs draw | Proposal standard deviation required |



# 14 Fit the spatial models using above R code

## 14.1 Fit exponential spatial model
```{r, message=FALSE,warning=FALSE}

# FIT EXPONENTIAL SPATIAL MODEL 

set.seed(54321)


priors_default <- list(
  beta_mean = 0,
  beta_var = 1000,
  sigma2_shape = 2,
  sigma2_rate = 1,
  tau2_shape = 2,
  tau2_rate = 1,
  phi_shape = 2,
  phi_rate = 1
)

mcmc_exp <- run_spatial_mcmc(
  data = litho2,
  response = "log_Thickness",
  covariates = c(
    "Surf_Elevation_scaled",
    "AB_Elevation_scaled"
  ),
  coords = litho2[, c("Easting", "Northing")],
  n_iter = 10000,
  burnin = 5000,
  thin = 5,
  cov_family = "exponential",
  priors = priors_default,
  initial_values = list(
    beta = NULL,
    W = NULL,
    sigma2 = 0.5,
    tau2 = 0.2,
    phi = 1
  ),
  phi_proposal_sd = 0.15,
  progress_every = 1000
)
```

## 14.2 Fit Matern spatial model

```{r, message=FALSE,warning=FALSE}

# Matérn covariance with smoothness kappa = nu = 1.5. The run_spatial_mcmc()
# function fixes nu at 1.5 and estimates sigma2, tau2 and phi.

mcmc_matern <- run_spatial_mcmc(
  data = litho2,
  response = "log_Thickness",
  covariates = c(
    "Surf_Elevation_scaled",
    "AB_Elevation_scaled"
  ),
  coords = litho2[, c("Easting", "Northing")],
  n_iter = 10000,
  burnin = 5000,
  thin = 5,
  cov_family = "matern",
  priors = priors_default,
  initial_values = list(
    beta = NULL,
    W = NULL,
    sigma2 = 0.5,
    tau2 = 0.2,
    phi = 1
  ),
  phi_proposal_sd = 0.15,
  progress_every = 1000
)

```

## 14.3  Function for posterior summary and diagnosie

```{r, message=FALSE, warnging=FALSE}

# POSTERIOR SUMMARY AND DIAGNOSTICS 

summarise_mcmc_model <- function(fit, model_name) {
  posterior_matrix <- as.matrix(fit$samples)

  posterior_summary <- t(apply(
    posterior_matrix,
    2,
    function(x) {
      c(
        Mean = mean(x),
        SD = sd(x),
        Median = median(x),
        `2.5%` = unname(quantile(x, 0.025)),
        `97.5%` = unname(quantile(x, 0.975))
      )
    }
  ))

  cat("\n===", toupper(model_name), "POSTERIOR SUMMARY ===\n")
  print(round(posterior_summary, 4))

  cat(
    "\n", model_name, "phi acceptance rate:",
    round(fit$acceptance_rate, 3), "\n"
  )

  if (fit$acceptance_rate < 0.20) {
    cat("Acceptance is low: consider reducing phi_proposal_sd.\n")
  } else if (fit$acceptance_rate > 0.50) {
    cat("Acceptance is high: consider increasing phi_proposal_sd.\n")
  } else {
    cat("The phi acceptance rate is in a reasonable range.\n")
  }

  # These diagnostics are protected so that a degenerate chain cannot
  # destroy the fitted model object.
  effective_sample_sizes <- tryCatch(
    coda::effectiveSize(fit$samples),
    error = function(e) {
      warning(model_name, " effectiveSize failed: ", conditionMessage(e))
      setNames(rep(NA_real_, ncol(posterior_matrix)), colnames(posterior_matrix))
    }
  )

  cat("\n===", toupper(model_name), "EFFECTIVE SAMPLE SIZES ===\n")
  print(effective_sample_sizes)

  autocorrelation <- tryCatch(
    coda::autocorr.diag(fit$samples, lags = c(1, 5, 10, 20)),
    error = function(e) {
      warning(model_name, " autocorr.diag failed: ", conditionMessage(e))
      NULL
    }
  )

  if (!is.null(autocorrelation)) {
    cat("\n===", toupper(model_name), "AUTOCORRELATIONS ===\n")
    print(round(autocorrelation, 3))
  }

  list(
    posterior_matrix = posterior_matrix,
    posterior_summary = posterior_summary,
    effective_sample_sizes = effective_sample_sizes,
    autocorrelation = autocorrelation
  )
}

```

## 14.4 Results for exponential spatial model

```{r, message=FALSE, warning=FALSE}

options(scipen = 999)
diagnostics_exp <- summarise_mcmc_model(mcmc_exp, "Exponential")
diagnostics_exp
plot( mcmc_exp$samples[, c( "beta_0", "beta_1", "beta_2" )])
plot( mcmc_exp$samples[, c(  "sigma2", "tau2", "phi")])


```

## 14.5 Results for Matern spatial model

```{r, message=FALSE, warning=FALSE}
options(scipen = 999)
diagnostics_matern <- summarise_mcmc_model(mcmc_matern, "Matern nu=1.5")
diagnostics_matern
plot( mcmc_matern$samples[, c( "beta_0", "beta_1", "beta_2" )])
plot( mcmc_matern$samples[, c(  "sigma2", "tau2", "phi")])

```

# 15 Use R package spBayes to fit the Bayesian spatial regression model

In fact, we do not need to write our own R function to conduct the Bayesian spatial regression models. Below we use spBayes::spLM() to run the analysis.


## 15.1 Prepare coordinates

The coordinates should be rescaled because large coordinate values can make the spatial decay parameter difficult to estimate.


```{r, message=FALSE,warning=FALSE}

# PREPARE COORDINATES ----------------------------------------------
library(tidyverse)
library(spBayes)
library(coda)
# Convert coordinates to kilometres if the original units are metres
coords <- as.matrix(
  litho2[, c("Easting", "Northing")]
) / 1000

colnames(coords) <- c("Easting_km", "Northing_km")

# Maximum pairwise distance
max_distance <- max(spBayes::iDist(coords))

max_distance

```

In **spBayes**, the exponential correlation function is parameterised as

$$R_{ij} = \exp(-\phi d_{ij}).$$

Therefore, $\phi$ is a spatial decay parameter. This differs from your custom function, which used

$$R_{ij} = \exp\left(-\frac{\phi}{\text{range}} d_{ij}\right).$$

The two parameters are related by

$$\phi_{\text{spBayes}} = \frac{1}{\phi_{\text{range}}}.$$

For the exponential covariance, the practical spatial range—where correlation is approximately 0.05—is

$$d_{0.05} \approx \frac{3}{\phi}.$$

## 15.2 Priors and starting values

`spLM()` uses:

- $\sigma^2 \sim \text{IG}(a_\sigma, b_\sigma)$
- $\tau^2 \sim \text{IG}(a_\tau, b_\tau)$

and a **Uniform** prior for $\phi$.


```{r, message=FALSE, warning=FALSE}
# PRIOR AND MCMC SETTINGS ------------------------------------------

response_variance <- var(litho2$log_Thickness)

# Phi bounds based on the maximum distance
phi_lower <- 3 / max_distance
phi_upper <- 300 / max_distance

# Starting value corresponding to an effective range
# of approximately half the maximum distance
phi_start <- 3 / (0.5 * max_distance)

starting_common <- list(
  sigma.sq = 0.7 * response_variance,
  tau.sq = 0.3 * response_variance,
  phi = phi_start
)

tuning_common <- list(
  sigma.sq = 0.05,
  tau.sq = 0.05,
  phi = 0.05
)

priors_common <- list(
  sigma.sq.ig = c(0.1, 0.1),
  tau.sq.ig = c(0.1, 0.1),
  phi.unif = c(phi_lower, phi_upper)
)

starting_common
priors_common


```

Unlike THE custom model, spLM() does not provide a Gamma prior for $\phi$; it uses a Uniform prior for the spatial decay parameter. The inverse-Gamma priors for the two variance components remain the same.


## 15.3 Exponential covariance model

```{r, message=FALSE, warning=FALSE}
# FIT EXPONENTIAL COVARIANCE MODEL ---------------------------------

set.seed(2026)

fit_exponential <- spLM(
  log_Thickness ~
    Surf_Elevation_scaled +
    AB_Elevation_scaled,

  data = litho2,
  coords = coords,

  starting = starting_common,
  tuning = tuning_common,
  priors = priors_common,

  cov.model = "exponential",

  n.samples = 30000,
  verbose = TRUE,
  n.report = 1000
)
```

The fitted object initially contains samples for the covariance parameters:

```{r}
names(fit_exponential)
summary(fit_exponential$p.theta.samples)

```

Recover the regression coefficients and spatial random effects after discarding the first 10,000 iterations and retaining every tenth sample:

```{r, message=FALSE, warning=FALSE}
fit_exponential <- spRecover(
  fit_exponential,
  start = 10001,
  end = 30000,
  thin = 10,
  get.beta = TRUE,
  get.w = TRUE,
  verbose = TRUE,
  n.report = 500
)

```

## 15.4 Matérn covariance model

For the Matérn model, fix the smoothness parameter at

$$\nu=1.5$$.

Setting its tuning variance to zero tells spLM() not to update it.

```{r, message=FALSE, warning=FALSE}
# 6. FIT MATERN COVARIANCE MODEL --------------------------------------

starting_matern <- c(
  starting_common,
  list(nu = 1.5)
)

tuning_matern <- c(
  tuning_common,
  list(nu = 0)
)

priors_matern <- c(
  priors_common,
  list(nu.unif = c(0.5, 2.5))
)

set.seed(2026)

fit_matern <- spLM(
  log_Thickness ~
    Surf_Elevation_scaled +
    AB_Elevation_scaled,

  data = litho2,
  coords = coords,

  starting = starting_matern,
  tuning = tuning_matern,
  priors = priors_matern,

  cov.model = "matern",

  n.samples = 30000,
  verbose = TRUE,
  n.report = 1000
)

fit_matern <- spRecover(
  fit_matern,
  start = 10001,
  end = 30000,
  thin = 10,
  get.beta = TRUE,
  get.w = TRUE,
  verbose = TRUE,
  n.report = 500
)
```

## 15.5 Posterior summaries

```{r, message=FALSE, warning=FALSE}
# 7. POSTERIOR SUMMARIES ----------------------------------------------

cat("\nExponential covariance parameters\n")

summary(
  window(
    fit_exponential$p.theta.samples,
    start = 10001,
    thin = 10
  )
)

cat("\nExponential regression coefficients\n")

summary(
  fit_exponential$p.beta.recover.samples
)


cat("\nMatérn covariance parameters\n")

summary(
  window(
    fit_matern$p.theta.samples,
    start = 10001,
    thin = 10
  )
)

cat("\nMatérn regression coefficients\n")

summary(
  fit_matern$p.beta.recover.samples
)
```

# 15.6 Create a concise posterior summary table:

```{r, message=FALSE, warning=FALSE}
posterior_summary <- function(samples, model_name) {

  samples <- as.matrix(samples)

  tibble(
    Model = model_name,
    Parameter = colnames(samples),
    Mean = apply(samples, 2, mean),
    SD = apply(samples, 2, sd),
    Median = apply(samples, 2, median),
    Lower_95 = apply(
      samples,
      2,
      quantile,
      probs = 0.025
    ),
    Upper_95 = apply(
      samples,
      2,
      quantile,
      probs = 0.975
    )
  )
}

beta_summary <- bind_rows(
  posterior_summary(
    fit_exponential$p.beta.recover.samples,
    "Exponential"
  ),
  posterior_summary(
    fit_matern$p.beta.recover.samples,
    "Matern"
  )
)

beta_summary
```
 
 
### 15.6.1 Resuts from the customer MCMC function of exponential spatial model

```{r,message=FALSE,warning=FALSE}
diagnostics_exp$posterior_summary

```
### 15.6.2 Resuts from the customer MCMC function of Matern spatial model 
 

```{r,message=FALSE,warning=FALSE}
diagnostics_matern$posterior_summary

```

## 15.7 MCMC diagnostic plots

Each parameter will have its own trace plot and posterior density plot. The model name should be used as the overall title, not as the name of every parameter.

```{r, message=FALSE, warning=FALSE}
# 8. MCMC DIAGNOSTICS -------------------------------------------------

theta_exp <- window(
  fit_exponential$p.theta.samples,
  start = 10001,
  thin = 10
)

theta_mat <- window(
  fit_matern$p.theta.samples,
  start = 10001,
  thin = 10
)

# Exponential covariance parameters
plot(
  theta_exp
  )

# Exponential regression coefficients
plot(
  fit_exponential$p.beta.recover.samples
)

# Matérn covariance parameters
par(mar = c(2, 2, 2, 2))
plot(
  theta_mat
  )


# Matérn regression coefficients
plot(
  fit_matern$p.beta.recover.samples
 )
```

## 15.8 Additional numerical diagnostics

```{r, message=FALSE, warning=FALSE}
# Effective sample sizes
effectiveSize(theta_exp)
effectiveSize(theta_mat)

effectiveSize(
  fit_exponential$p.beta.recover.samples
)

effectiveSize(
  fit_matern$p.beta.recover.samples
)

# Geweke convergence diagnostics
geweke.diag(theta_exp)
geweke.diag(theta_mat)

geweke.diag(
  fit_exponential$p.beta.recover.samples
)

geweke.diag(
  fit_matern$p.beta.recover.samples
)

```

## 15.9 Convert $\phi$ into practical spatial range
For the exponential model:
```{r, message=FALSE,warning=FALSE}
phi_exp <- as.matrix(theta_exp)[, "phi"]

effective_range_exp <- 3 / phi_exp

quantile(
  effective_range_exp,
  probs = c(0.025, 0.5, 0.975)
)
```

This is now a package-based analysis: spLM() performs the MCMC estimation, while  spRecover() generates posterior samples for $\beta$ and W(s). The current spBayes documentation confirms that spLM() supports exponential and Matérn covariance models, inverse-Gamma priors for $\sigma^2$ and $\tau^2$, a Uniform prior for phi, and a flat default prior for the regression coefficients. CRAN spLM documentation





